### 2025-01-31 products/manuscript/Manuscript.qmd
Part One Completed
Part One components - description of data set, research question formation, ideas for data cleaning all included in this commit-- also officially obtained my 2019-2024 human tick-bourne illness case data in conjunction with my internship that I am doing! Super excited to proceed.

---
title: "Analysis of Tick-Bourne Disease Cases Reported to DPH Between 2019-2024"
author: ""
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../../assets/dataanalysis-references.bib
csl: ../../assets/american-journal-of-epidemiology.csl
---

The structure below is one possible setup for a manuscript, or a general data analysis project (including the course project). Adjust as needed. You don't need to have exactly these sections, but the content covering those sections should be addressed.

This uses MS Word as output format. [See here](https://quarto.org/docs/output-formats/ms-word.html) for more information. You can switch to other formats, like html or pdf. See [the Quarto documentation](https://quarto.org/) for other formats.

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(knitr)
```

**Authors**

-   Hope Grismer $^{1,2,*, \land}$ (ORCID: 0000-0000-1234-5678)
-   Second Author$^{3, *}$
-   Third Author$^{3, \dagger}$
-   Last Author$^{1, 2,}$

**Author affiliations**

1.  College of Public Health, University of Georgia, Athens, GA, USA.
2.  Southeastern Center of Excellence in Vector-Bourne Diseases (SECVBD), Gainesville, FL 32611.
3.  Yet another Affiliation.

$*$ These authors contributed equally to this work.

$\land$ Corresponding author: hope.grismer\@uga.edu

$\dagger$ Disclaimer: The opinions expressed in this article are the author's own and don't reflect their employer.

{{< pagebreak >}}

# Summary/Abstract

*Write a summary of your project.*

{{< pagebreak >}}

# Introduction

## General Background Information

Tick-borne illnesses are diseases that humans can contract from the bite of an infected tick. Ticks are small, blood-sucking arachnids that are often found in wooded, grassy, or suburban areas. When a tick feeds on an infected animal, it can acquire pathogens—such as bacteria, viruses, or parasites—that can then be transmitted to humans. Examples of tick-bourne diseases include, but are not limited to, Lyme disease, anaplasmosis, and babesiosis. These diseases are of increasing public health concern, especially as the geographic range and seasonality of ticks expand due to environmental and ecological changes.

## Description of data and data source

This data set includes all human cases reported to the Georgia Department of Public Health (GA-DPH) between 2019 and 2024. It was graciously provided by DPH professionals in the epidemiology department. All available data for each case over this five-year period across the state of Georgia. Proper de-identification practices-- the removal of first name, last name, and date of birth (DOB)-- has been taken. The dataset includes variables such as disease type, patient demographics (gender, age, district, state, county, and ZIP code), medical details (incident ID, report ID, lab test dates, results, serotype, and titers), as well as disease outcomes (death status, fever, migranes, and administrative status). Reported diagnoses included in this dataset include Toxoplasmosis, Dengue, Rocky Mountain Spotted Fever, Anaplasma Phagocytophilum, Malaria, Lyme Disease, West Nile (WNV) Infection, Chikungunya, LaCrosse (LAC) Infection, Zika, Ehrlichia Chaffeensis, and Typhus (Epidemic/Louse/Flea). Some of this cases, based on adherence to specific case definitions and subsequent lab results, are categorized as suspect, probable, confirmed, or unconfirmed. By examining case reports, public health professionals can better understand the prevalence, demographics, and regional variations of these diseases. This analysis aims to further inform the public health understanding of tick-bourne disease in Georgia.

## Questions/Hypotheses to be addressed

1.  How have vector-borne diseases (VBD) changed over time, and what seasonal patterns or climatic factors influence these trends?

2.  What is the distribution of vector-borne diseases across different health districts, counties, and ZIP codes?

3.  How complete is the data for vector-borne diseases, and how has data completeness evolved over time?

4.  What is the timeline between key dates (onset, lab test, and date of entry), excluding same-day onset and lab results, and how does this timeline vary over time?

5.  What trends exist in malaria travel cases over time, and which countries are most frequently visited by affected individuals?

6.  What are the trends and changes observed in relative incidence of different disease pathologies between 2019-2024?

To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above and have the right bibtex key. Then you can include like this:

Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].

{{< pagebreak >}}

# Methods

***I. Describing the Uncleaned Data***

The data contains 6255 entries representing discrete cases of tick-bourne diseases. For each case, some or all of the following data variables were collected: disease type, patient demographics (gender, age, district, state, county, and ZIP code), medical details (incident ID, report ID, lab test dates, results, serotype, and titers), and disease outcomes (death status, fever, migranes, and administrative status). For some cases, specific variables were not reported and, therefore, are reported as blank spaces in the data set.

***II. Initial Data Cleaning Stages***

Primary data cleaning anticipated at this initial stage (this section to be rewritten later)

-   Remove all cases that were "NOT CONFIRMED," "UNCONFIRMED," or left blank in the STATUS category

-   Remove entries that contain any blanks to ensure all data analyzed later include complete profiles

-   Sort by Gender, County, and Age

***III. Analysis Process***

-   Still determining this

## Schematic of workflow

Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it). @fig-schematic is an example of some - completely random/unrelated - schematic that was generated with Biorender. We store those figures in the `assets` folder.

```{r}
#| label: fig-schematic
#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
#| echo: FALSE
knitr::include_graphics(here("assets","antigen-recognition.png"))
```

## Data aquisition

*As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next.*

## Data import and cleaning

*Write code that reads in the file and cleans it so it's ready for analysis. Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts. If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g. as supplement in a paper). All materials, including files that contain code, should be commented well so everyone can follow along.*

## Statistical analysis

*Explain anything related to your statistical analyses.*

{{< pagebreak >}}

# Results

## Exploratory/Descriptive analysis

*Use a combination of text/tables/figures to explore and describe your data. Show the most important descriptive results here. Additional ones should go in the supplement. Even more can be in the R and Quarto files that are part of your project.*

@tbl-summarytable shows a summary of the data.

Note the loading of the data providing a **relative** path using the `../../` notation. (Two dots means a folder up). You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path. You can also use the `here` R package to create paths. See examples of that below. I generally recommend the `here` package.

```{r}
#| label: tbl-summarytable
#| tbl-cap: "Data summary table."
#| echo: FALSE
resulttable=readRDS("../../results/tables/summarytable.rds")
knitr::kable(resulttable)
```

## Basic statistical analysis

*To get some further insight into your data, if reasonable you could compute simple statistics (e.g. simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable. Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*

@fig-result shows a scatterplot figure produced by one of the R scripts.

```{r}
#| label: fig-result
#| fig-cap: "Height and weight stratified by gender."
#| echo: FALSE
knitr::include_graphics(here("results","figures","height-weight-stratified.png"))
```

## Full analysis

*Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc. This might again be code that is best placed in one or several separate R scripts that need to be well documented. You want the code to produce figures and data ready for display as tables, and save those. Then you load them here.*

Example @tbl-resulttable2 shows a summary of a linear model fit.

```{r}
#| label: tbl-resulttable2
#| tbl-cap: "Linear model fit table."
#| echo: FALSE
resulttable2 = readRDS(here("results","tables","resulttable2.rds"))
knitr::kable(resulttable2)
```

{{< pagebreak >}}

# Discussion

## Summary and Interpretation

*Summarize what you did, what you found and what it means.*

## Strengths and Limitations

*Discuss what you perceive as strengths and limitations of your analysis.*

## Conclusions

*What are the main take-home messages?*

*Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end*

This paper [@leek2015] discusses types of analyses.

These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.

Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header. Many more style files for almost any journal [are available](https://www.zotero.org/styles). You also specify the location of your bibtex reference file in the YAML. You can call your reference file anything you like.

{{< pagebreak >}}

# References

### 2025-01-31 code/processing-code/processingcode2.R

@@ -1,0 +1,75 @@
+###############################
+# processing script
+#
+#this script loads the raw data, processes and cleans it 
+#and saves it as Rds file in the processed-data folder
+#
+# Note the ## ---- name ---- notation
+# This is done so one can pull in the chunks of code into the Quarto document
+# see here: https://bookdown.org/yihui/rmarkdown-cookbook/read-chunk.html
+###############################
+
+
+## ---- packages --------
+#load needed packages. make sure they are installed.
+library(readxl) #for loading Excel files
+library(dplyr) #for data processing/cleaning
+library(tidyr) #for data processing/cleaning
+library(skimr) #for nice visualization of data 
+library(here) #to set paths
+
+
+## ---- loaddata --------
+#path to data
+#note the use of the here() package and not absolute paths
+data_location <- here::here("data","raw-data","Tick-Bourne_Cases_2019_to_2024.xlsx")
+#load data. 
+#note that for functions that come from specific packages (instead of base R)
+# I often specify both package and function like so
+#package::function() that's not required one could just call the function
+#specifying the package makes it clearer where the function "lives",
+#but it adds typing. You can do it either way.
+rawdata <- readxl::read_excel(data_location)
+
+## ---- exploredata --------
+#take a look at the data
+dplyr::glimpse(rawdata)
+#another way to look at the data
+summary(rawdata)
+#yet another way to get an idea of the data
+head(rawdata)
+#this is a nice way to look at data
+skimr::skim(rawdata)
+
+#data cleaning step one 
+
+
+
+## ---- savedata --------
+# all done, data is clean now. 
+# Let's assign at the end to some final variable
+# makes it easier to add steps above
+processeddata2 <- d4
+# location to save file
+save_data_location <- here::here("data","processed-data","processeddata2.rds")
+saveRDS(processeddata2, file = save_data_location)
+
+
+## ---- notes --------
+# anything you don't want loaded into the Quarto file but 
+# keep in the R file, just give it its own label and then don't include that label
+# in the Quarto file
+
+# Dealing with NA or "bad" data:
+# removing anyone who had "faulty" or missing data is one approach.
+# it's often not the best. based on your question and your analysis approach,
+# you might want to do cleaning differently (e.g. keep individuals with some missing information)
+
+# Saving data as RDS:
+# I suggest you save your processed and cleaned data as RDS or RDA/Rdata files. 
+# This preserves coding like factors, characters, numeric, etc. 
[…]

### 2025-02-21 code/processing-code/processing-code-tickcases.qmd
Part Two Completed
Part Two is completed I believe; Note to self on reference manager and double checking my ReadMe files

---
title: "Data Cleaning for Raw Tick-Bourne Cases Data - 2019 to 2024"
author: "Hope Grismer"
date: "2025-02-21"
output: html_document
---

Data Cleaning for Raw Tick-Bourne Cases Data - 2019 to 2024

```{r}
library(readxl)  # For loading Excel files
library(dplyr)   # For data processing/cleaning
library(tidyr)   # For data processing/cleaning
library(skimr)   # For nice visualization of data
library(here)    # To set paths
library(kableExtra)
```

```{r}
## ---- loaddata --------
# Define the path to the data
data_location <- here::here("data", "raw-data", "Tick-Bourne_Cases_2019_to_2024.xlsx")
```

```{r}
# Load the data
rawdata <- readxl::read_excel(data_location)
```

```{r}
## ---- exploredata --------
# Take a look at the raw data structure
dplyr::glimpse(rawdata)
head(rawdata)
```

```{r}
# Summary statistics
summary(rawdata)

# Preview first few rows
head(rawdata)

# Nice visualization of data
skimr::skim(rawdata)
```
The first cleaning step removes cases where the STATUS field is either missing or not labeled as "CONFIRMED." This ensures that only officially confirmed cases are included in further analysis. The cleaned dataset is then saved for future reference.

```{r}
## ---- cleandata --------
# Remove cases where STATUS is NA or not confirmed
cleaned_data <- rawdata %>%
  dplyr::filter(!is.na(STATUS) & STATUS == "CONFIRMED")

# Take another look at cleaned data
dplyr::glimpse(cleaned_data)

## ---- savedata --------
# Define save location for cleaned data
save_data_location <- here::here("data", "processed-data", "Tick-Bourne_Cases_Cleaned.rds")

# Ensure the "processed-data" folder exists
if (!dir.exists(here::here("data", "processed-data"))) {
  dir.create(here::here("data", "processed-data"), recursive = TRUE)
}
# Save the cleaned data
saveRDS(cleaned_data, file = save_data_location)

# Confirm success
message("Cleaned data saved successfully at: ", save_data_location)
```

```{r}
## ---- Table --------
cleaned_data %>%
  dplyr::select(DISEASE, AGE, GENDER, COUNTY, STATE, LABDATE1, RESULT1, STATUS) %>%  # Select key columns for display
  knitr::kable(caption = "Cleaned Tick-Bourne Disease Cases (2019-2024)", align = "c") %>%
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"),
    full_width = FALSE,
    position = "center"
  ) %>%
  kableExtra::column_spec(1, bold = TRUE, color = "white", background = "#2C3E50") %>%  # Highlight the first column
  kableExtra::column_spec(2:8, width = "12em") %>%  # Adjust column widths for readability
  kableExtra::row_spec(0, bold = TRUE, background = "#34495E", color = "white")  # Style header row
```
To refine the dataset further, only cases with a confirmed "positive" test result are retained. Any cases with missing, blank, or ambiguous results (such as "inconclusive") are excluded. This step eliminated 15 cases that did not meet the case definition used in Georgia, leaving 95 valid data points. However, upon further review, I noticed that additional positive cases that were not explicitly labeled "positive" were inadvertently removed, prompting a refinement of the filtering criteria.

```{r}
## ---- clean_data_step2 --------
# Further exclude cases where RESULT1 is blank or not "positive" in any capitalization
cleaned_data2 <- cleaned_data %>%
  filter(!is.na(RESULT1) & RESULT1 != "" & tolower(RESULT1) == "positive")

# Save final cleaned dataset
final_save_data_location <- here::here("data", "processed-data", "Tick-Bourne_Cases_Cleaned_Step2.rds")
saveRDS(cleaned_data2, file = final_save_data_location)

## ---- display_table_step2 --------
# Display professional table for second cleaning step
kableExtra::kable(cleaned_data2, format = "html", caption = "Final Cleaned Data After Removing Blank RESULT1 Cases") %>%
  kableExtra::kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  kableExtra::column_spec(1, bold = TRUE, color = "white", background = "#2C3E50") %>%  # Highlight first column
  kableExtra::column_spec(2:ncol(cleaned_data2), width = "10em") %>%  # Adjust column widths
  kableExtra::row_spec(0, bold = TRUE, background = "#34495E", color = "white")  # Header row styling
```
This step provided an easier summary table of some of the most relevant information on these positive cases.

```{r}
## ---- display only a few of the columns --------

# Select the specified columns and sort by DISEASE
selected_columns_data <- cleaned_data2 %>%
  select(DISEASE, STATUS, COUNTY, GENDER, AGE, LABTEST1, RESULT1) %>%
  dplyr::arrange(DISEASE)  # Ensure sorting by disease name

# Display the table with consistent styling
kableExtra::kable(selected_columns_data, format = "html", caption = "Selected Columns from Final Cleaned Data") %>%
  kableExtra::kable_styling(
    full_width = FALSE, 
    bootstrap_options = c("striped", "hover", "condensed", "responsive")
  ) %>%
  kableExtra::column_spec(1, bold = TRUE, color = "white", background = "#2C3E50") %>%  # Highlight the first column
  kableExtra::column_spec(1:ncol(selected_columns_data), width = "12em") %>%  # Adjust column widths
  kableExtra::row_spec(0, bold = TRUE, background = "#2C3E50", color = "white")  # Header row styling
```
A second round of filtering ensures that all positive cases are correctly identified while excluding known non-positive values such as "Negative," "Equivocal," "Not detected," and other irrelevant entries. The updated dataset is saved again for consistency.
```{r}
## ---- clean_data_step3--------
# Define values to exclude from RESULT1
excluded_values <- c("", NA, "Negative", "Equivocal", "negative", "Choose One", "Not detected", "Indeterminant/Equivocal")

# Filter out only the excluded values while keeping all other results
cleaned_data3 <- cleaned_data %>%
  filter(!(RESULT1 %in% excluded_values) & !is.na(RESULT1))

# Save final cleaned dataset
cleaned_data3_save_location <- here::here("data", "processed-data", "Tick-Bourne_Cases_Final_Cleaned.rds") 
saveRDS(cleaned_data3, file = cleaned_data3_save_location)

# Display count of remaining observations
nrow(cleaned_data3)

```
To present the cleaned data effectively, selected key columns—including DISEASE, STATUS, COUNTY, GENDER, AGE, LABTEST1, and RESULT1—are displayed in a professionally styled table using the kableExtra package. This enhances readability and makes it easier to analyze trends.
```{r}
## ---- display only a few of the columns --------

# Select the specified columns
selected_columns_data <- cleaned_data3 %>%
  select(DISEASE, STATUS, COUNTY, GENDER, AGE, LABTEST1, RESULT1) %>%
  dplyr::arrange(DISEASE)  # Use the correct column name

# Display the table with consistent styling
kableExtra::kable(selected_columns_data, format = "html", caption = "Selected Columns from Final Cleaned Data") %>%
  kableExtra::kable_styling(
    full_width = FALSE, 
    bootstrap_options = c("striped", "hover", "condensed", "responsive")
  ) %>%
  kableExtra::column_spec(1, bold = TRUE, color = "white", background = "#2C3E50") %>%  # Highlight the first column
  kableExtra::column_spec(1:ncol(selected_columns_data), width = "12em") %>%  # Adjust column widths
  kableExtra::row_spec(0, bold = TRUE, background = "#2C3E50", color = "white")  # Header row styling

```
A histogram is created to visualize the distribution of different tick-borne diseases in Georgia between 2019 and 2024. The ggplot2 package is used to generate a clean and informative bar chart, illustrating the most prevalent diseases. 
```{r}
## ---- disease_histogram --------
library(ggplot2)

# Create histogram of disease counts
disease_plot <- ggplot(cleaned_data3, aes(x = DISEASE)) +
  geom_bar(fill = "#2C3E50", color = "white") +  # Dark blue bars with white edges
  theme_minimal() +
  labs(
    title = "Distribution of Tick-Bourne Disease\nTypes for Confirmed Cases in Georgia (2019-2024)",
    x = "Disease Type",
    y = "Count"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),  # Rotate & reduce x-axis label size
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    plot.margin = margin(10, 10, 20, 10)  # Adjust margins to prevent cutoff
  )

# Display the plot
print(disease_plot)

# Define save locations
save_plot_location <- here::here("results", "figures", "disease_histogram.png")

# Ensure directory exists
dir.create(here::here("results", "figures"), recursive = TRUE, showWarnings = FALSE)

# Save the plot
ggsave(filename = save_plot_location, plot = disease_plot, width = 12, height = 6, dpi = 300)
```
To explore the severity of Rocky Mountain Spotted Fever cases, a box plot of TITER1 values (log-transformed for better visualization) is created. This helps in understanding the spread and variability of test results among confirmed cases. It also is considered relatively analagous to the patterns seen in Alpha-Gal Syndrome, which is a topic of interest for me, but is not explicitly included in this data set as it is currently not a reportable disease in Georgia.
```{r}
# Save the plot
ggsave(filename = save_plot_location, plot = disease_plot, width = 12, height = 6, dpi = 300)

# ---- rmsf_boxplot --------
# Filter for Rocky Mountain Spotted Fever cases
rmsf_data <- rawdata %>%
  filter(DISEASE == "ROCKY MOUNTAIN SPOTTED FEVER") %>%
  mutate(TITER1 = as.numeric(TITER1)) %>%  # Ensure TITER1 is numeric
  filter(!is.na(TITER1) & TITER1 > 0)

# Create box and whisker plot for log-transformed TITER1 values
rmsf_titer_plot <- ggplot(rmsf_data, aes(y = TITER1)) +
  geom_boxplot(fill = "#1B9E77", color = "black") +  # Green box plot
  scale_y_log10() +  # Apply log10 scale
  theme_minimal() +
  labs(
    title = "Log-Transformed Distribution of TITER1 Values\nfor Rocky Mountain Spotted Fever Cases",
    y = "TITER1 Value (Log Scale)"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    axis.title.x = element_blank(),  # Remove x-axis title
    axis.text.x = element_blank(),   # Remove x-axis labels
    axis.ticks.x = element_blank()   # Remove x-axis ticks
  )

# Display the plot
print(rmsf_titer_plot)

# Define save location
save_rmsf_plot_location <- here::here("results", "figures", "rmsf_titer_boxplot_log.png")

# Save the plot
ggsave(filename = save_rmsf_plot_location, plot = rmsf_titer_plot, width = 8, height = 6, dpi = 300)

```
The dataset is filtered to determine the five most common tick-borne diseases. A bar chart is created to highlight the case count for each of these diseases, providing an overview of the most significant public health concerns.
```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(readr)
library(here)

# Load the cleaned data
cleaned_data <- readRDS(here::here("data", "processed-data", "Tick-Bourne_Cases_Cleaned.rds"))

# Get top 5 diseases by total count across all counties
top_diseases <- cleaned_data %>%
  count(DISEASE, sort = TRUE) %>%  # Count cases per disease
  top_n(5, n) %>%  # Select top 5 diseases
  pull(DISEASE)  # Extract disease names as a vector

# Filter data for only the top 5 diseases
filtered_data <- cleaned_data %>%
  filter(DISEASE %in% top_diseases)

# Summarize total cases for each of the top 5 diseases
disease_summary <- filtered_data %>%
  group_by(DISEASE) %>%
  summarise(Count = n(), .groups = "drop")  # Count cases per Disease

# Create bar chart
top5_disease_plot <- ggplot(disease_summary, aes(x = reorder(DISEASE, -Count), y = Count, fill = DISEASE)) +
  geom_bar(stat = "identity", color = "black", alpha = 0.8) +  # Bar chart for count
  geom_text(aes(label = Count), vjust = -0.5, size = 5) +  # Count labels above bars
  theme_minimal() +
  labs(
    title = "Top 5 Tick-Borne Diseases",
    x = "Disease",
    y = "Case Count",
    fill = "Disease"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10)  # Angled labels for clarity
  ) +
  scale_fill_brewer(palette = "Set1")  # Nice color scheme for different diseases

# Display the plot
print(top5_disease_plot)

# Define save location
save_top5_disease_plot <- here::here("results", "figures", "top5_diseases_plot.png")

# Save the plot
ggsave(filename = save_top5_disease_plot, plot = top5_disease_plot, width = 8, height = 6, dpi = 300)

# Confirm success
message("Plot saved successfully at: ", save_top5_disease_plot)

```
Lyme Disease is likewise similar to Alpha-gal. So, a specific analysis is conducted for Lyme Disease cases, focusing on test results was conducted. Below ia a count histogram generated to compare positive, negative, and other test outcomes, offering insights into testing patterns and diagnosis rates.

```{r}
# ---- lyme_count_histogram --------
# Filter for Lyme Disease cases and exclude "Choose One" category
lyme_data2 <- rawdata %>%
  filter(DISEASE == "LYME DISEASE", RESULT1 != "Choose One") %>%
  mutate(
    RESULT1 = case_when(
      RESULT1 %in% c("Unknown", "Other") ~ "Unknown/Other",  # Group "Unknown" and "Other"
      RESULT1 == "Results Pending" ~ "Results Pending",      # Keep "Results Pending" separate
      TRUE ~ RESULT1  # Keep all other values unchanged
    ),
    RESULT1 = factor(RESULT1)  # Ensure it's a factor
  )

# Create count histogram for positive vs. negative cases
lyme_count_histogram2 <- ggplot(lyme_data2, aes(x = RESULT1, fill = RESULT1)) +
  geom_bar(color = "black", alpha = 0.8) +  # Bar chart for count comparison
  geom_text(stat = "count", aes(label = after_stat(count)), vjust = -0.5, size = 5) +  # Count labels above bars
  scale_fill_manual(values = c(
    "Positive" = "#E41A1C",
    "Negative" = "#377EB8",
    "Unknown/Other" = "#4DAF4A",   # Green for the grouped category
    "Results Pending" = "#FF7F00"  # Orange for "Results Pending"
  )) +
  theme_minimal() +
  labs(
    title = "Count of Lyme Disease Test Results",
    x = "Test Result",
    y = "Count",
    fill = "Test Result"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    axis.text.x = element_text(face = "bold", size = 8, angle = 45, hjust = 1)  # Smaller labels
  )

# Display the plot
print(lyme_count_histogram2)

# Define save location
save_lyme_count_plot <- here::here("results", "figures", "lyme_result_count_histogram2.png")

# Save the plot
ggsave(filename = save_lyme_count_plot, plot = lyme_count_histogram2, width = 8, height = 6, dpi = 300)

```

The final visualization maps tick-borne disease cases across different counties in Georgia. A grouped bar chart displays case counts per county, categorized by disease type. This geographic breakdown helps in identifying regional hotspots for tick-borne illnesses, which can inform public health interventions. This visualization is busy so it was further altered to show the top five tick bourne diseases and how they break down across the top 12 (by overall count of confirmed cases) counties in Georgia.

```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(readr)
library(here)

# Load the cleaned data
cleaned_data <- readRDS(here::here("data", "processed-data", "Tick-Bourne_Cases_Cleaned.rds"))

# Create count summary by Disease and County
disease_county_summary <- cleaned_data %>%
  group_by(DISEASE, COUNTY) %>%
  summarise(Count = n(), .groups = "drop")  # Count cases per Disease & County

# Create bar chart
disease_county_plot <- ggplot(disease_county_summary, aes(x = COUNTY, y = Count, fill = DISEASE)) +
  geom_bar(stat = "identity", color = "black", alpha = 0.8) +  # Bar chart for count
  geom_text(aes(label = Count), vjust = -0.5, size = 3) +  # Count labels above bars
  theme_minimal() +
  labs(
    title = "Tick-Borne Disease Cases by County",
    x = "County",
    y = "Case Count",
    fill = "Disease"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8)  # Angled labels for clarity
  ) +
  scale_fill_brewer(palette = "Set1")  # Nice color scheme for different diseases

# Display the plot
print(disease_county_plot)

# Define save location
save_disease_county_plot <- here::here("results", "figures", "disease_county_plot.png")

# Save the plot
ggsave(filename = save_disease_county_plot, plot = disease_county_plot, width = 10, height = 6, dpi = 300)

# Confirm success
message("Plot saved successfully at: ", save_disease_county_plot)

```
To look at trends in disease burden as well as "hot-spot" counties, I generated the stacked bar graph below.
```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(readr)
library(here)
library(RColorBrewer)

# Load the cleaned data
cleaned_data <- readRDS(here::here("data", "processed-data", "Tick-Bourne_Cases_Cleaned.rds"))

# Get top 5 diseases by total count across all counties
top_diseases <- cleaned_data %>%
  count(DISEASE, sort = TRUE) %>%  
  top_n(5, n) %>%  
  pull(DISEASE)

# Filter data for only the top 5 diseases
filtered_data <- cleaned_data %>%
  filter(DISEASE %in% top_diseases)

# Identify the top 12 counties by total case count
top_counties <- filtered_data %>%
  count(COUNTY, sort = TRUE) %>%
  top_n(12, n) %>%
  pull(COUNTY)

# Ensure we have exactly 12 counties (or fewer if dataset is small)
top_counties <- unique(top_counties)[1:min(12, length(top_counties))]

# Keep only those top counties in the dataset
filtered_data <- filtered_data %>%
  filter(COUNTY %in% top_counties)

# Summarize case counts for each disease and county
disease_county_summary <- filtered_data %>%
  group_by(DISEASE, COUNTY) %>%
  summarise(Count = n(), .groups = "drop")

# Get the exact number of unique counties in filtered data
num_counties <- length(unique(disease_county_summary$COUNTY))

# Assign colors dynamically based on available counties
county_colors <- setNames(brewer.pal(num_counties, "Paired"), unique(disease_county_summary$COUNTY))

# Create stacked bar chart
top5_stacked_plot <- ggplot(disease_county_summary, aes(x = DISEASE, y = Count, fill = COUNTY)) +
  geom_bar(stat = "identity", color = "black", alpha = 0.8) +  
  theme_minimal() +
  labs(
    title = "Top 5 Tick-Borne Diseases by County",
    x = "Disease",
    y = "Case Count",
    fill = "County"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 14),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10)
  ) +
  scale_fill_manual(values = county_colors)  # Assign fixed colors to the top counties

# Display the plot
print(top5_stacked_plot)

# Define save location
save_top5_stacked_plot <- here::here("results", "figures", "top5_diseases_stacked_plot.png")

# Save the plot
ggsave(filename = save_top5_stacked_plot, plot = top5_stacked_plot, width = 8, height = 6, dpi = 300)

# Confirm success
message("Plot saved successfully at: ", save_top5_stacked_plot)

```
Age distribution might be of interest in the analysis, so this was checked as part of this initial data cleaning. This output shows 0 rows of the cleaned data to be missing ages. As an exploratory analysis, I looked at disease count by age group from my cleaned data set. 
```{r}
missing_age_rows <- cleaned_data3 %>%
  filter(is.na(AGE))

print(missing_age_rows)

library(ggplot2)
library(dplyr)

# Create age bins of 20 years
selected_columns_data <- cleaned_data2 %>%
  select(DISEASE, AGE) %>%
  mutate(Age_Bin = cut(AGE, 
                       breaks = seq(0, 100, by = 20),  # Adjust max age as needed
                       labels = c("0-19", "20-39", "40-59", "60-79", "80+"),
                       right = FALSE)) %>%
  drop_na(Age_Bin)  # Remove any NA values due to missing AGE data

# Create stacked bar plot
age_group_plot <- ggplot(selected_columns_data, aes(x = Age_Bin, fill = DISEASE)) +
  geom_bar(position = "stack") +
  labs(title = "Disease Count by Age Group",
       x = "Age Group",
       y = "Count",
       fill = "Disease Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels for readability

# Save the plot to results/figures using here::here()
ggsave(
  filename = here::here("results", "figures", "disease_count_by_age.png"), 
  plot = age_group_plot, 
  width = 8, 
  height = 6, 
  dpi = 300
)

age_group_plot

```
Another thing that might be used in my analysis to answer the question of how data completeness evolved over time is cleaning the data to extract data onset information.

By distinguishing between Blank and Text in the STATUS column over time, this plot helps pinpoint periods with high missingness. If there are spikes in Blank values at certain times, it could indicate data entry issues, system errors, or gaps in reporting that need to be addressed before deeper analysis.
```{r}
library(ggplot2)
library(dplyr)
library(lubridate)

# Convert DOO to Date and group by Month
statuscompleteness_summary <- rawdata %>%
  mutate(DOO = as.Date(DOO),  # Ensure DOO is in Date format
         Month = floor_date(DOO, "month"),  # Aggregate by month
         Status_Category = ifelse(STATUS == "" | is.na(STATUS), "Blank", "Text")) %>%
  group_by(Month, Status_Category) %>%
  summarise(Count = n(), .groups = "drop")

# Save the cleaned dataset to the processed_data folder
write_csv(statuscompleteness_summary, here::here("data", "processed-data", "cleaned_statuscompleteness_summary.csv"))

# Create improved line plot
status_completeness_plot <- ggplot(statuscompleteness_summary, aes(x = Month, y = Count, color = Status_Category, group = Status_Category)) +
  geom_line(size = 1) +  # Adjust line thickness
  geom_point(size = 1, alpha = 0.6) +  # Reduce point size & add transparency
  geom_smooth(method = "loess", se = FALSE, linetype = "dashed") +  # Add smoothed trendline
  labs(title = "Comparison of Blank vs. Text in STATUS Column by Month",
       x = "Date of Onset (Monthly Aggregation)",
       y = "Count",
       color = "Status Category") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels

# Save the plot to results/figures using here::here()
ggsave(
  filename = here::here("results", "figures", "status_completeness_plot.png"), 
  plot = status_completeness_plot, 
  width = 8, 
  height = 6, 
  dpi = 300
)
status_completeness_plot

```

```{r}
library(dplyr)
library(lubridate)
library(readr)
library(here)
library(kableExtra)
library(fs)

# Extract year and month from the selected date columns
processed_date_data <- rawdata %>%
  mutate(
    UPDATE_YEAR = year(ymd(UPDATEDATE)),
    UPDATE_MONTH = month(ymd(UPDATEDATE), label = TRUE),  # Extract month name
    DOO_YEAR = year(ymd(DOO)),
    DOO_MONTH = month(ymd(DOO), label = TRUE),
    LAB1_YEAR = year(ymd(LABDATE1)),
    LAB1_MONTH = month(ymd(LABDATE1), label = TRUE),
    LAB2_YEAR = year(ymd(LABDATE2)),
    LAB2_MONTH = month(ymd(LABDATE2), label = TRUE)
  ) %>%
  filter(
    !is.na(UPDATE_YEAR) & !is.na(UPDATE_MONTH) & 
    !is.na(DOO_YEAR) & !is.na(DOO_MONTH) & 
    !is.na(LAB1_YEAR) & !is.na(LAB1_MONTH) & 
    !is.na(LAB2_YEAR) & !is.na(LAB2_MONTH)  # Exclude rows with missing date parts
  )

# Select key columns for review
summary_table <- processed_date_data %>%
  select(DISEASE, UPDATEDATE, DOO, LABDATE1, LABDATE2)

# Save the cleaned dataset
fs::dir_create(here::here("data", "processed-data"))  # Ensure directory exists
write_csv(processed_date_data, here::here("data", "processed-data", "processed_dates.csv"))

# Summary Table
kableExtra::kable(summary_table, format = "html", caption = "Extracted Date Information for Disease Cases") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  kableExtra::column_spec(1, bold = TRUE, color = "white", background = "#2C3E50") %>%  # Highlight the first column
  kableExtra::column_spec(2:ncol(summary_table), width = "12em") %>%  # Adjust column widths
  kableExtra::row_spec(0, bold = TRUE, background = "#2C3E50", color = "white")  # Header row styling

```
Another question that I want to analyze later and therefore needed to clean for was What is the timeline between key dates (onset, lab test, and date of entry), excluding same-day onset and lab results, and how does this timeline vary over time? One thing I have noticed in my internship work on tick-bourne illness is the huge issue with loss-to-follow-up.

The code below calculate the time differences between key dates. The script then calculates the number of days between the onset date (DOO) and the first lab date (LABDATE1), between the first and second lab dates (LABDATE2), and between the second lab date and the update date (UPDATEDATE). These values are summed to determine the total duration of each case.

This processed dataset is then saved as a CSV file for further analysis. The script also generates a formatted HTML table to display the extracted time information in a structured manner.

Once I took a look at this table, however, there were big issues. Namely, that negative time (in days) was recorded between the initial lab result and the follow up test. I wrote further code to remove these values because it is not able to be determined what the proper date was due to user error. I added a filter accordingly.

```{r}
library(dplyr)
library(lubridate)
library(readr)
library(here)
library(kableExtra)
library(fs)
library(ggplot2)

# Calculate time differences between the dates
processed_time_data <- processed_date_data %>%
  mutate(
    DOO_to_LAB1 = as.numeric(difftime(LABDATE1, DOO, units = "days")),
    LAB1_to_LAB2 = as.numeric(difftime(LABDATE2, LABDATE1, units = "days")),
    LAB2_to_UPDATE = as.numeric(difftime(UPDATEDATE, LABDATE2, units = "days")),
    TOTAL_DAYS = DOO_to_LAB1 + LAB1_to_LAB2 + LAB2_to_UPDATE
  )

# Remove rows with any negative values
processed_time_data2 <- processed_time_data %>%
  filter(if_all(c(DOO_to_LAB1, LAB1_to_LAB2, LAB2_to_UPDATE, TOTAL_DAYS), ~ . >= 0))

# Select key columns for review
time_summary_table <- processed_time_data2 %>%
  select(DISEASE, DOO, LABDATE1, LABDATE2, UPDATEDATE, DOO_to_LAB1, LAB1_to_LAB2, LAB2_to_UPDATE, TOTAL_DAYS)

# Save the dataset with time differences
fs::dir_create(here::here("data", "processed-data"))  # Ensure directory exists
write_csv(processed_time_data2, here::here("data", "processed-data", "time_between_processeddata.csv"))

# Summary Table
kableExtra::kable(time_summary_table, format = "html", caption = "Time Differences Between Key Date Points") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  kableExtra::column_spec(1, bold = TRUE, color = "white", background = "#2C3E50") %>%  # Highlight the first column
  kableExtra::column_spec(2:ncol(time_summary_table), width = "12em") %>%  # Adjust column widths
  kableExtra::row_spec(0, bold = TRUE, background = "#2C3E50", color = "white")  # Header row styling

```

### 2025-02-21 products/manuscript/Manuscript.qmd
Part Two Completed
Part Two is completed I believe; Note to self on reference manager and double checking my ReadMe files

@@
 csl: ../../assets/american-journal-of-epidemiology.csl
+editor: 
+  markdown: 
+    wrap: 72
 ---
 
-The structure below is one possible setup for a manuscript, or a general data analysis project (including the course project).
-Adjust as needed.
-You don't need to have exactly these sections, but the content covering those sections should be addressed.
+The structure below is one possible setup for a manuscript, or a general
+data analysis project (including the course project). Adjust as needed.
+You don't need to have exactly these sections, but the content covering
+those sections should be addressed.
 
-This uses MS Word as output format.
-[See here](https://quarto.org/docs/output-formats/ms-word.html) for more information.
-You can switch to other formats, like html or pdf.
-See [the Quarto documentation](https://quarto.org/) for other formats.
+This uses MS Word as output format. [See
+here](https://quarto.org/docs/output-formats/ms-word.html) for more
+information. You can switch to other formats, like html or pdf. See [the
+Quarto documentation](https://quarto.org/) for other formats.
 
@@
 1.  College of Public Health, University of Georgia, Athens, GA, USA.
-2.  Southeastern Center of Excellence in Vector-Bourne Diseases (SECVBD), Gainesville, FL 32611.
+2.  Southeastern Center of Excellence in Vector-Bourne Diseases
+    (SECVBD), Gainesville, FL 32611.
 3.  Yet another Affiliation.
@@
 
-$\dagger$ Disclaimer: The opinions expressed in this article are the author's own and don't reflect their employer.
+$\dagger$ Disclaimer: The opinions expressed in this article are the
+author's own and don't reflect their employer.
 
@@
 
-Tick-borne illnesses are diseases that humans can contract from the bite of an infected tick.
-Ticks are small, blood-sucking arachnids that are often found in wooded, grassy, or suburban areas.
-When a tick feeds on an infected animal, it can acquire pathogens—such as bacteria, viruses, or parasites—that can then be transmitted to humans.
-Examples of tick-bourne diseases include, but are not limited to, Lyme disease, anaplasmosis, and babesiosis.
-These diseases are of increasing public health concern, especially as the geographic range and seasonality of ticks expand due to environmental and ecological changes.
+Tick-borne illnesses are diseases that humans can contract from the bite
+of an infected tick. Ticks are small, blood-sucking arachnids that are
+often found in wooded, grassy, or suburban areas. When a tick feeds on
+an infected animal, it can acquire pathogens—such as bacteria, viruses,
+or parasites—that can then be transmitted to humans. Examples of
+tick-bourne diseases include, but are not limited to, Lyme disease,
+anaplasmosis, and babesiosis. These diseases are of increasing public
+health concern, especially as the geographic range and seasonality of
+ticks expand due to environmental and ecological changes.
 
@@
 
-This data set includes all human cases reported to the Georgia Department of Public Health (GA-DPH) between 2019 and 2024.
-It was graciously provided by DPH professionals in the epidemiology department.
-All available data for each case over this five-year period across the state of Georgia.
-Proper de-identification practices-- the removal of first name, last name, and date of birth (DOB)-- has been taken.
-The dataset includes variables such as disease type, patient demographics (gender, age, district, state, county, and ZIP code), medical details (incident ID, report ID, lab test dates, results, serotype, and titers), as well as disease outcomes (death status, fever, migranes, and administrative status).
-Reported diagnoses included in this dataset include Toxoplasmosis, Dengue, Rocky Mountain Spotted Fever, Anaplasma Phagocytophilum, Malaria, Lyme Disease, West Nile (WNV) Infection, Chikungunya, LaCrosse (LAC) Infection, Zika, Ehrlichia Chaffeensis, and Typhus (Epidemic/Louse/Flea).
-Some of this cases, based on adherence to specific case definitions and subsequent lab results, are categorized as suspect, probable, confirmed, or unconfirmed.
-By examining case reports, public health professionals can better understand the prevalence, demographics, and regional variations of these diseases.
-This analysis aims to further inform the public health understanding of tick-bourne disease in Georgia.
+This data set includes all human cases reported to the Georgia
+Department of Public Health (GA-DPH) between 2019 and 2024. It was
+graciously provided by DPH professionals in the epidemiology department.
+All available data for each case over this five-year period across the
+state of Georgia. Proper de-identification practices-- the removal of
+first name, last name, and date of birth (DOB)-- has been taken. The
+dataset includes variables such as disease type, patient demographics
+(gender, age, district, state, county, and ZIP code), medical details
+(incident ID, report ID, lab test dates, results, serotype, and titers),
+as well as disease outcomes (death status, fever, migranes, and
+administrative status). Reported diagnoses included in this dataset
+include Toxoplasmosis, Dengue, Rocky Mountain Spotted Fever, Anaplasma
+Phagocytophilum, Malaria, Lyme Disease, West Nile (WNV) Infection,
+Chikungunya, LaCrosse (LAC) Infection, Zika, Ehrlichia Chaffeensis, and
+Typhus (Epidemic/Louse/Flea). Some of this cases, based on adherence to
+specific case definitions and subsequent lab results, are categorized as
+suspect, probable, confirmed, or unconfirmed. By examining case reports,
+public health professionals can better understand the prevalence,
+demographics, and regional variations of these diseases. This analysis
+aims to further inform the public health understanding of tick-bourne
+disease in Georgia.
 
@@
 
-1.  How have vector-borne diseases (VBD) changed over time, and what seasonal patterns or climatic factors influence these trends?
+1.  How have vector-borne diseases (VBD) changed over time, and what
+    seasonal patterns or climatic factors influence these trends?
 
-2.  What is the distribution of vector-borne diseases across different health districts, counties, and ZIP codes?
+2.  What is the distribution of vector-borne diseases across different
+    health districts, counties, and ZIP codes?
 
-3.  How complete is the data for vector-borne diseases, and how has data completeness evolved over time?
+3.  How complete is the data for vector-borne diseases, and how has data
+    completeness evolved over time?
 
-4.  What is the timeline between key dates (onset, lab test, and date of entry), excluding same-day onset and lab results, and how does this timeline vary over time?
+4.  What is the timeline between key dates (onset, lab test, and date of
+    entry), excluding same-day onset and lab results, and how does this
+    timeline vary over time?
 
-5.  What trends exist in malaria travel cases over time, and which countries are most frequently visited by affected individuals?
+5.  What trends exist in malaria travel cases over time, and which
+    countries are most frequently visited by affected individuals?
 
-6.  What are the trends and changes observed in relative incidence of different disease pathologies between 2019-2024?
+6.  What are the trends and changes observed in relative incidence of
+    different disease pathologies between 2019-2024?
 
-To cite other work (important everywhere, but likely happens first in introduction), make sure your references are in the bibtex file specified in the YAML header above and have the right bibtex key.
-Then you can include like this:
+To cite other work (important everywhere, but likely happens first in
+introduction), make sure your references are in the bibtex file
+specified in the YAML header above and have the right bibtex key. Then
+you can include like this:
 
-Examples of reproducible research projects can for instance be found in [@mckay2020; @mckay2020a].
+Examples of reproducible research projects can for instance be found in
+[@mckay2020; @mckay2020a].
 
@@
 
-***I. Describing the Uncleaned Data***
+**I. Uncleaned Data**
 
-The data contains 6255 entries representing discrete cases of tick-bourne diseases.
-For each case, some or all of the following data variables were collected: disease type, patient demographics (gender, age, district, state, county, and ZIP code), medical details (incident ID, report ID, lab test dates, results, serotype, and titers), and disease outcomes (death status, fever, migranes, and administrative status).
-For some cases, specific variables were not reported and, therefore, are reported as blank spaces in the data set.
+The data contains 6255 entries representing discrete cases of
+tick-bourne diseases. For each case, some or all of the following data
+variables were collected: disease type, patient demographics (gender,
+age, district, state, county, and ZIP code), medical details (incident
+ID, report ID, lab test dates, results, serotype, and titers), and
+disease outcomes (death status, fever, migranes, and administrative
+status). For some cases, specific variables were not reported and,
+therefore, are reported as blank spaces in the data set.
 
@@
 
-Primary data cleaning anticipated at this initial stage (this section to be rewritten later)
+Looked for Missing Data & General Trends
 
--   Remove all cases that were "NOT CONFIRMED," "UNCONFIRMED," or left blank in the STATUS category
+-   Identified Disease Burden Trends & Hotspot Counties
 
--   Remove entries that contain any blanks to ensure all data analyzed later include complete profiles
+<!-- -->
 
--   Sort by Gender, County, and Age
+-   Generated a stacked bar graph to visualize case distribution across
+    regions. - Checked Age Distribution
+
+<!-- -->
+
+-   Verified that no cases had missing age values in the cleaned
+    dataset.
+
+<!-- -->
+
+-   Conducted an exploratory analysis of disease counts by age group.
+
+-   Evaluated Data Completeness Over Time
+
+-   Extracted onset date information to track missing data trends.
+
+-   Created a plot distinguishing between blank and non-blank STATUS
+    entries over time.
+
+-   Identified periods of high missingness, which may indicate data
+    entry errors or reporting gaps.
+
+Analyzed Timeline Between Key Case Events
+
+-   Calculated time differences for:
+
+    -   Onset date (DOO) to first lab test date (LABDATE1)
+
+    -   First lab test (LABDATE1) to second lab test (LABDATE2)
+
+    -   Second lab test (LABDATE2) to case update date (UPDATEDATE)
+
+    -   Summed these values to determine the total case duration. -
+        Saved the processed dataset as a CSV file for further
+        analysis. - Created a formatted HTML table to present extracted
+        time intervals.
+
+Addressed Data Quality Issues
+
+-   Identified negative time values in lab test dates, indicating data
+    entry errors
+
+-   Implemented a filtering step to remove records with invalid date
+    sequences
 
@@
 
-Sometimes you might want to show a schematic diagram/figure that was not created with code (if you can do it with code, do it).
-@fig-schematic is an example of some - completely random/unrelated - schematic that was generated with Biorender.
-We store those figures in the `assets` folder.
-
-```{r}
-#| label: fig-schematic
-#| fig-cap: "A figure that is manually generated and shows some overview/schematic. This has nothing to do with the data, it's just a random one from one of our projects I found and placed here."
-#| echo: FALSE
-knitr::include_graphics(here("assets","antigen-recognition.png"))
-```
+Sometimes you might want to show a schematic diagram/figure that was not
+created with code (if you can do it with code, do it). @fig-schematic is
+an example of some - completely random/unrelated - schematic that was
+generated with Biorender. We store those figures in the `assets` folder.
 
@@
 
-*As applicable, explain where and how you got the data. If you directly import the data from an online source, you can combine this section with the next.*
+The data used in this analysis was obtained through my internship with
+the Georgia Department of Health, in collaboration with the Centers for
+Disease Control and Prevention (CDC). Specifically, it is derived from
+the Nationally Notifiable Diseases Surveillance System (NNDSS) and
+focuses on tick-borne illnesses. This dataset provides key insights into
+the trends, distribution, and characteristics of reported cases,
+supporting public health efforts in disease monitoring and prevention.
 
@@
 
-*Write code that reads in the file and cleans it so it's ready for analysis.
-Since this will be fairly long code for most datasets, it might be a good idea to have it in one or several R scripts.
-If that is the case, explain here briefly what kind of cleaning/processing you do, and provide more details and well documented code somewhere (e.g.
-as supplement in a paper).
-All materials, including files that contain code, should be commented well so everyone can follow along.*
+The data cleaning and preprocessing workflow begins with loading
+essential R packages for data handling (dplyr, tidyr, readxl), summary
+visualization (skimr), and file path management (here). The raw dataset,
+stored in an Excel file, is imported into R for structured processing.
+To assess data quality and structure, we utilize functions such as
+glimpse(), head(), and summary(), which provide an overview of column
+names, data types, and potential inconsistencies like missing or
+unexpected values. Additionally, skimr::skim() offers a detailed
+summary, highlighting numerical distributions and missing data patterns.
+
+The initial cleaning step removes records where the STATUS field is
+either missing or not classified as "CONFIRMED," ensuring that only
+officially verified cases are included. The dataset is then refined
+further by retaining only cases with a confirmed "positive" test result,
+excluding records with missing, blank, or ambiguous values (e.g.,
+"inconclusive"). This process initially removed 15 cases that did not
+meet the case definition in Georgia, leaving 95 valid records. Upon
+further review, filtering criteria were adjusted to ensure all relevant
+positive cases were retained while systematically excluding non-positive
+values such as "Negative," "Equivocal," "Not detected," and other
+irrelevant categories. This refinement resulted in a more accurate
+dataset and a clearer summary table of key case information. The cleaned
+dataset is saved at each stage to maintain version control and
+reproducibility.
+
+## Key Variables in Data Wrangling - Preparing for Analysis
+
+To effectively present the cleaned dataset, key variables—including
+DISEASE, STATUS, COUNTY, GENDER, AGE, LABTEST1, and RESULT1—are
+displayed in a professionally formatted table using the kableExtra
+package. This enhances readability and facilitates trend analysis.
+Several visualizations are generated using ggplot2 to explore patterns
+in tick-borne diseases in Georgia between 2019 and 2024: Disease
+Distribution: A histogram illustrates the prevalence of different
+tick-borne diseases, highlighting the most commonly reported cases.
+
+**Rocky Mountain Spotted Fever (RMSF) Severity:**
+
+A box plot of TITER1 values (log-transformed for better visualization)
+provides insights into the variability of test results. RMSF severity
+patterns are considered analogous to Alpha-Gal Syndrome, a topic of
+interest, though not explicitly included in this dataset as it is not a
+reportable disease in Georgia.
+
+```{r}
+library(here)
+library(knitr)
+
+# Generate correct file path
+img_path <- here::here("results", "figures", "rmsf_titer_boxplot_log.png")
+
+# Display the image
+knitr::include_graphics(img_path)
+
+```
+
+**Most Common Tick-Borne Diseases:**
+
+The dataset is filtered to identify the five most frequently reported
+tick-borne diseases. A bar chart visualizes their case counts, offering
+an overview of significant public health concerns.
+
+```{r}
+library(here)
+library(knitr)
+
+# Generate correct file path
+img_path <- here::here("results", "figures", "disease_histogram.png")
+
+# Display the image
+knitr::include_graphics(img_path)
+```
+
+**Lyme Disease Analysis:**
+
+Given its similarity to Alpha-Gal Syndrome
+(which is of interest to me and the topic of my capstone), Lyme Disease
+cases are analyzed separately. A count histogram compares positive,
+negative, and other test outcomes, providing insights into testing
+patterns and diagnosis rates.
+
+```{r}
+library(here)
+library(knitr)
+
+# Generate correct file path
+img_path <- here::here("results", "figures", "lyme_result_count_histogram2.png")
+
+# Display the image
+knitr::include_graphics(img_path)
+
+```
+
+**Geographic Distribution:**
+
+A grouped bar chart maps case counts across
+Georgia counties, categorized by disease type. Due to the density of
+information, this visualization is refined to focus on the top five
+tick-borne diseases and their distribution across the 12 counties with
+the highest confirmed case counts.
+```{r}
+library(here)
+library(knitr)
+
+# Generate correct file paths
+img_paths <- c(
+  here::here("results", "figures", "disease_county_plot.png"),
+  here::here("results", "figures", "top5_disease_county_plot.png")
+)
+
+# Display the images
+knitr::include_graphics(img_paths)
+```
+
+To analyze disease burden trends and identify hotspot counties, a
+stacked bar graph is generated to visualize case distribution across
+different regions.
+
+**Age Distribution Analysis**
+
+Age distribution is assessed as part of
+the initial data cleaning process, confirming that no cases in the
+cleaned dataset have missing age values. Additionally, an exploratory
+analysis examines disease counts by age group to identify potential
+patterns in disease prevalence.
+```{r}
+library(here)
+library(knitr)
+
+# Generate correct file path
+img_path <- here::here("results", "figures", "disease_count_by_age.png")
+
+# Display the image
+knitr::include_graphics(img_path)
+```
+
+**Data Completeness Over Time** 
+To evaluate how data completeness has
+evolved, the dataset is processed to extract onset date information. A
+visualization distinguishes between blank and non-blank entries in the
+STATUS column over time, helping to identify periods with high
+missingness. Spikes in blank values may indicate data entry errors,
+system issues, or reporting gaps that should be addressed before
+conducting further analysis.
+
+```{r}
+library(here)
+library(knitr)
+
+# Generate correct file path
+img_path <- here::here("results", "figures", "status_completeness_plot.png")
+
+# Display the image
+knitr::include_graphics(img_path)
+```
+
+**Timeline Analysis of Key Case Events** Given the frequent
+loss-to-follow-up observed in tick-borne illness reporting, an analysis
+is conducted to examine time gaps between critical dates, including:
+
+**Onset date (DOO) to first lab test date (LABDATE1)**
+
+**First lab test (LABDATE1) to second lab test (LABDATE2)**
+
+**Second lab test (LABDATE2) to case update date (UPDATEDATE)**
+
+These durations are summed to determine the total case timeline for each
+record. The processed dataset is saved as a CSV file for further
+analysis, and a formatted HTML table is generated to present the
+extracted time intervals in a structured manner. Data Quality
+Adjustments Upon reviewing the time calculations, inconsistencies were
+identified where negative time values appeared between initial lab
+results and follow-up tests. These errors, likely due to data entry
+mistakes, made it impossible to determine the correct dates. To address
+this, a filtering step was added to remove erroneous records where
+follow-up test dates occurred before initial lab dates.
 
@@
 
-*Use a combination of text/tables/figures to explore and describe your data.
-Show the most important descriptive results here.
-Additional ones should go in the supplement.
-Even more can be in the R and Quarto files that are part of your project.*
+*Use a combination of text/tables/figures to explore and describe your
+data. Show the most important descriptive results here. Additional ones
+should go in the supplement. Even more can be in the R and Quarto files
+that are part of your project.*
 
-@tbl-summarytable shows a summary of the data.
-
-Note the loading of the data providing a **relative** path using the `../../` notation.
-(Two dots means a folder up).
-You never want to specify an **absolute** path like `C:\ahandel\myproject\results\` because if you share this with someone, it won't work for them since they don't have that path.
-You can also use the `here` R package to create paths.
-See examples of that below.
-I generally recommend the `here` package.
-
-```{r}
-#| label: tbl-summarytable
-#| tbl-cap: "Data summary table."
-#| echo: FALSE
-resulttable=readRDS("../../results/tables/summarytable.rds")
-knitr::kable(resulttable)
-```
+Note the loading of the data providing a **relative** path using the
+`../../` notation. (Two dots means a folder up). You never want to
+specify an **absolute** path like `C:\ahandel\myproject\results\`
+because if you share this with someone, it won't work for them since
+they don't have that path. You can also use the `here` R package to
+create paths. See examples of that below. I generally recommend the
+`here` package.
 
@@
 
-*To get some further insight into your data, if reasonable you could compute simple statistics (e.g.
-simple models with 1 predictor) to look for associations between your outcome(s) and each individual predictor variable.
-Though note that unless you pre-specified the outcome and main exposure, any "p\<0.05 means statistical significance" interpretation is not valid.*
-
-@fig-result shows a scatterplot figure produced by one of the R scripts.
-
-```{r}
-#| label: fig-result
-#| fig-cap: "Height and weight stratified by gender."
-#| echo: FALSE
-knitr::include_graphics(here("results","figures","height-weight-stratified.png"))
-```
-
 ## Full analysis
 
-*Use one or several suitable statistical/machine learning methods to analyze your data and to produce meaningful figures, tables, etc.
-This might again be code that is best placed in one or several separate R scripts that need to be well documented.
-You want the code to produce figures and data ready for display as tables, and save those.
-Then you load them here.*
-
-Example @tbl-resulttable2 shows a summary of a linear model fit.
-
-```{r}
-#| label: tbl-resulttable2
-#| tbl-cap: "Linear model fit table."
-#| echo: FALSE
-resulttable2 = readRDS(here("results","tables","resulttable2.rds"))
-knitr::kable(resulttable2)
-```
+*Use one or several suitable statistical/machine learning methods to
+analyze your data and to produce meaningful figures, tables, etc. This
+might again be code that is best placed in one or several separate R
+scripts that need to be well documented. You want the code to produce
+figures and data ready for display as tables, and save those. Then you
+load them here.*
 
@@
 
-*Discuss what you perceive as strengths and limitations of your analysis.*
+*Discuss what you perceive as strengths and limitations of your
+analysis.*
 
@@
 
-*Include citations in your Rmd file using bibtex, the list of references will automatically be placed at the end*
-
-This paper [@leek2015] discusses types of analyses.
-
-These papers [@mckay2020; @mckay2020a] are good examples of papers published using a fully reproducible setup similar to the one shown in this template.
-
-Note that this cited reference will show up at the end of the document, the reference formatting is determined by the CSL file specified in the YAML header.
-Many more style files for almost any journal [are available](https://www.zotero.org/styles).
-You also specify the location of your bibtex reference file in the YAML.
-You can call your reference file anything you like.
+*Include citations in your Rmd file using bibtex, the list of references
+will automatically be placed at the end*
 

### 2025-02-21 code/processing-code/processingcode2.R

@@ -1,75 +1,0 @@
-###############################
-# processing script
-#
-#this script loads the raw data, processes and cleans it 
-#and saves it as Rds file in the processed-data folder
-#
-# Note the ## ---- name ---- notation
-# This is done so one can pull in the chunks of code into the Quarto document
-# see here: https://bookdown.org/yihui/rmarkdown-cookbook/read-chunk.html
-###############################
-
-
-## ---- packages --------
-#load needed packages. make sure they are installed.
-library(readxl) #for loading Excel files
-library(dplyr) #for data processing/cleaning
-library(tidyr) #for data processing/cleaning
-library(skimr) #for nice visualization of data 
-library(here) #to set paths
-
-
-## ---- loaddata --------
-#path to data
-#note the use of the here() package and not absolute paths
-data_location <- here::here("data","raw-data","Tick-Bourne_Cases_2019_to_2024.xlsx")
-#load data. 
-#note that for functions that come from specific packages (instead of base R)
-# I often specify both package and function like so
-#package::function() that's not required one could just call the function
-#specifying the package makes it clearer where the function "lives",
-#but it adds typing. You can do it either way.
-rawdata <- readxl::read_excel(data_location)
-
-## ---- exploredata --------
-#take a look at the data
-dplyr::glimpse(rawdata)
-#another way to look at the data
-summary(rawdata)
-#yet another way to get an idea of the data
-head(rawdata)
-#this is a nice way to look at data
-skimr::skim(rawdata)
-
-#data cleaning step one 
-
-
-
-## ---- savedata --------
-# all done, data is clean now. 
-# Let's assign at the end to some final variable
-# makes it easier to add steps above
-processeddata2 <- d4
-# location to save file
-save_data_location <- here::here("data","processed-data","processeddata2.rds")
-saveRDS(processeddata2, file = save_data_location)
-
-
-## ---- notes --------
-# anything you don't want loaded into the Quarto file but 
-# keep in the R file, just give it its own label and then don't include that label
-# in the Quarto file
-
-# Dealing with NA or "bad" data:
-# removing anyone who had "faulty" or missing data is one approach.
-# it's often not the best. based on your question and your analysis approach,
-# you might want to do cleaning differently (e.g. keep individuals with some missing information)
-
-# Saving data as RDS:
-# I suggest you save your processed and cleaned data as RDS or RDA/Rdata files. 
-# This preserves coding like factors, characters, numeric, etc. 
[…]

### 2025-03-02 code/processing-code/processing-code-tickcases.qmd
Part Two Edits

@@
 ```
+
 The first cleaning step removes cases where the STATUS field is either missing or not labeled as "CONFIRMED." This ensures that only officially confirmed cases are included in further analysis.
@@
 ```
+
 To refine the dataset further, only cases with a confirmed "positive" test result are retained.
@@
 ```
+
 This step provided an easier summary table of some of the most relevant information on these positive cases.
@@
 ```
+
 A second round of filtering ensures that all positive cases are correctly identified while excluding known non-positive values such as "Negative," "Equivocal," "Not detected," and other irrelevant entries.
 The updated dataset is saved again for consistency.
+
 ```{r}
@@
 ```
+
 To present the cleaned data effectively, selected key columns—including DISEASE, STATUS, COUNTY, GENDER, AGE, LABTEST1, and RESULT1—are displayed in a professionally styled table using the kableExtra package.
 This enhances readability and makes it easier to analyze trends.
+
 ```{r}
@@
 ```
+
 A histogram is created to visualize the distribution of different tick-borne diseases in Georgia between 2019 and 2024.
 The ggplot2 package is used to generate a clean and informative bar chart, illustrating the most prevalent diseases.
+
 ```{r}
@@
 ```
+
 To explore the severity of Rocky Mountain Spotted Fever cases, a box plot of TITER1 values (log-transformed for better visualization) is created.
@@
 It also is considered relatively analagous to the patterns seen in Alpha-Gal Syndrome, which is a topic of interest for me, but is not explicitly included in this data set as it is currently not a reportable disease in Georgia.
+
 ```{r}
@@
 ```
+
 The dataset is filtered to determine the five most common tick-borne diseases.
 A bar chart is created to highlight the case count for each of these diseases, providing an overview of the most significant public health concerns.
+
 ```{r}
@@
 ```
+
 Lyme Disease is likewise similar to Alpha-gal.
@@
 ```
+
 To look at trends in disease burden as well as "hot-spot" counties, I generated the stacked bar graph below.
+
 ```{r}
@@
 ```
+
 Age distribution might be of interest in the analysis, so this was checked as part of this initial data cleaning.
@@
 As an exploratory analysis, I looked at disease count by age group from my cleaned data set.
+
 ```{r}
@@
 ```
+
 Another thing that might be used in my analysis to answer the question of how data completeness evolved over time is cleaning the data to extract data onset information.
@@
 If there are spikes in Blank values at certain times, it could indicate data entry issues, system errors, or gaps in reporting that need to be addressed before deeper analysis.
+
 ```{r}
@@
 ```
+
 Another question that I want to analyze later and therefore needed to clean for was What is the timeline between key dates (onset, lab test, and date of entry), excluding same-day onset and lab results, and how does this timeline vary over time?

### 2025-03-02 products/manuscript/Manuscript.qmd
Part Two Edits

@@
   here::here("results", "figures", "disease_county_plot.png"),
+  here::here("results", "figures", "top5_diseases_stacked_plot"),
   here::here("results", "figures", "top5_disease_county_plot.png")

### 2025-03-13 products/manuscript/Manuscript.qmd
Fixed rendering error
Found an error that I made after initial render check in omitting a .png that made it so the manuscript could not find the file that I created in the processing code

--- 2025-02-21 products/manuscript/Manuscript.qmd
@@
   here::here("results", "figures", "disease_county_plot.png"),
+  here::here("results", "figures", "top5_diseases_stacked_plot.png"),
   here::here("results", "figures", "top5_disease_county_plot.png")

### 2025-03-17 code/analysis-code/statistical-analysis.qmd
Part 3 Done
Made some better manuscript edits

---
title: "Data Analysis for Raw Tick-Bourne Cases Data - 2019 to 2024"
author: "Hope Grismer"
date: "2025-03-10"
output: html_document
---

######################################
# analysis script
#
# This script loads the processed, cleaned data, does a simple analysis
# and saves the results to the results folder.

First, load all the libraries needed for tidymodels, multivariable GLM, and data visualization.

```{r}
library(tidymodels)  # For tidymodels framework
library(ggplot2)     # For data visualization
library(dplyr)       # For data manipulation
library(lubridate)   # For date manipulation
library(readr)       # For reading and writing CSV files
library(fs)          # For file system operations
library(broom)       # For tidying up GLM results
library(here)        # for data loading/saving
```

Then, load the cleaned datasets.

```{r}
# Load the cleaned dataset
cleaned_data2 <- readRDS(here::here("data", "processed-data", "Tick-Bourne_Cases_Cleaned_Step2.rds"))
processed_time_data2 <- read_csv(here::here("data", "processed-data", "time_between_processeddata.csv"))

```

While the processing code creation involved some exploratory data analysis (EDA) already to help me determine what cleaning needed to be done, additional EDA helped find patterns in data to analyze further. See below.

```{r}

# Visualize time differences (e.g., DOO to LAB1)
ggplot(processed_time_data2, aes(x = DOO_to_LAB1)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Time from DOO to LAB1 (in days)", 
       x = "Days from DOO to LAB1", y = "Frequency")

```

######################################
# Data fitting/statistical analysis
#####################################

```{r}
# Path to the cleaned data (after saving it)
cleaned_data3 <- readRDS(here::here("data", "processed-data", "Tick-Bourne_Cases_Final_Cleaned.rds"))
```

```{r}
# 1. Disease trends over time:
# Extract YEAR from DOO (assuming DOO is a Date column)
cleaned_data3 <- cleaned_data3 %>%
  mutate(YEAR = as.integer(format(as.Date(DOO, format = "%Y-%m-%d"), "%Y")))

# Count cases grouped by DISEASE type and YEAR
disease_counts <- cleaned_data3 %>%
  group_by(DISEASE, YEAR) %>%
  summarise(CASE_COUNT = n(), .groups = "drop")

# Fit a linear model for yearly trend in cases (by DISEASE)
lmfit1 <- lm(CASE_COUNT ~ YEAR + DISEASE, data = disease_counts)
lmtable1 <- broom::tidy(lmfit1)
print(lmtable1)

# Save fit results
table_file1 <- here::here("results", "tables", "trend_over_time_by_disease.rds")
saveRDS(lmtable1, file = table_file1)

# Plot yearly trend by disease
plot1 <- ggplot(disease_counts, aes(x = YEAR, y = CASE_COUNT, color = DISEASE)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm") +
  labs(
    title = "Vector-Borne Diseases Over Time by Disease",
    x = "Year",
    y = "Number of Cases"
  ) +
  theme_minimal()

plot1

# Save plot
plot_file1 <- here::here("results", "figures", "trend_over_time_by_disease.png")
ggsave(plot_file1, plot = plot1, width = 8, height = 6, dpi = 300)
```

```{r}
# 2. Distribution Across Health Districts:
# Bar plot of cases by health district
plot2 <- ggplot(cleaned_data3, aes(x = COUNTY, fill = DISEASE)) +
  geom_bar() +
  labs(
    title = "Cases by Health District",
    x = "Health District",
    y = "Number of Cases"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plot2

# Save plot
plot_file2 <- here::here("results", "figures", "cases_by_health_district.png")
ggsave(plot_file2, plot = plot2, width = 8, height = 6, dpi = 300)
```

```{r}
# 3. Completeness 
library(ggplot2)
library(dplyr)
library(lubridate)
library(broom)  # for tidying model results
library(readr)  # for reading CSV files

# Load the cleaned data (adjust path as needed)
statuscompleteness_summary_location <- here::here("data", "processed-data", "cleaned_statuscompleteness_summary.csv")
statuscompleteness_summary <- read_csv(statuscompleteness_summary_location)

# 1. Data Completeness Analysis Over Time:
# Summarize completeness over time
completeness_summary <- statuscompleteness_summary %>%
  mutate(YEAR = year(Month)) %>%
  group_by(YEAR) %>%
  summarise(COMPLETENESS = mean(Count) / sum(Count) * 100, .groups = "drop")  # Calculate completeness as percentage

# Linear Model: Analyze the relationship between YEAR and COMPLETENESS
lm_completeness <- lm(COMPLETENESS ~ YEAR, data = completeness_summary)

# Summarize the linear model results
lm_completeness_summary <- broom::tidy(lm_completeness)
print(lm_completeness_summary)

# 2. Create the completeness plot with linear trendline
plot3 <- ggplot(completeness_summary, aes(x = YEAR, y = COMPLETENESS)) +
  geom_line(color = "darkgreen") +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red", linetype = "dashed") +  # Add linear model trendline
  labs(
    title = "Data Completeness Over Time",
    x = "Year",
    y = "Completeness (%)"
  ) +
  theme_minimal()

# Print the plot
plot3

# Save the plot to the results/figures folder
plot_file3 <- here::here("results", "figures", "completeness_over_time.png")
ggsave(plot_file3, plot = plot3, width = 8, height = 6, dpi = 300)

```

```{r}
# 4. Timeline Between Key Dates:
# Fit a model for days between onset and lab over time
filtered_time_data <- cleaned_data3 %>%
  filter(!is.na(LABDATE1) & !is.na(DOO) & (LABDATE1 > DOO))

# Calculate the days between onset and lab result (assuming DATEONSET and LABDATE1 are in Date format)
filtered_time_data <- filtered_time_data %>%
  mutate(DAYS_TO_LAB = as.integer(LABDATE1 - DOO))

lmfit2 <- lm(DAYS_TO_LAB ~ YEAR, data = filtered_time_data)
lmtable2 <- broom::tidy(lmfit2)
print(lmtable2)

# Save fit results
table_file2 <- here::here("results", "tables", "days_to_lab_over_time.rds")
saveRDS(lmtable2, file = table_file2)

# Plot days between onset and lab over time
plot4 <- ggplot(filtered_time_data, aes(x = YEAR, y = DAYS_TO_LAB)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", color = "red") +
  labs(
    title = "Days Between Onset and Lab Processing Over Time",
    x = "Year",
    y = "Days Between Onset and Lab"
  ) +
  theme_minimal()

plot4

# Save plot
plot_file4 <- here::here("results", "figures", "days_to_lab_over_time.png")
ggsave(plot_file4, plot = plot4, width = 8, height = 6, dpi = 300)

```

```{r}
# 5. Relative Incidence of Disease Pathologies (2019–2024):
# Filter data for years 2019–2024
incidence_data <- cleaned_data3 %>%
  filter(YEAR >= 2019 & YEAR <= 2024)

# Group and calculate incidence by disease
incidence_summary <- incidence_data %>%
  group_by(YEAR, DISEASE) %>%
  summarise(INCIDENCE = n()) %>%
  ungroup()

# Plot relative incidence over time
plot5 <- ggplot(incidence_summary, aes(x = YEAR, y = INCIDENCE, color = DISEASE)) +
  geom_line(size = 1) +
  labs(
    title = "Relative Incidence of Disease Pathologies (2019–2024)",
    x = "Year",
    y = "Incidence"
  ) +
  theme_minimal()

plot5

# Save plot
plot_file5 <- here::here("results", "figures", "relative_incidence_2019_2024.png")
ggsave(plot_file5, plot = plot5, width = 8, height = 6, dpi = 300)
```

### 2025-03-17 code/processing-code/processing-code-tickcases.qmd
Part 3 Done
Made some better manuscript edits

@@
 
-Another question that I want to analyze later and therefore needed to clean for was What is the timeline between key dates (onset, lab test, and date of entry), excluding same-day onset and lab results, and how does this timeline vary over time?
-One thing I have noticed in my internship work on tick-bourne illness is the huge issue with loss-to-follow-up.
+Another question that I want to analyze later and therefore needed to clean for was "what is the timeline between key dates (onset, lab test, and date of entry), excluding same-day onset and lab results, and how does this timeline vary over time?" One thing I have noticed in my internship work on tick-bourne illness is the huge issue with loss-to-follow-up.
 

### 2025-03-17 products/manuscript/Manuscript.qmd
Part 3 Done
Made some better manuscript edits

@@
 tick-bourne diseases include, but are not limited to, Lyme disease,
-anaplasmosis, and babesiosis. These diseases are of increasing public
+Anaplasmosis, and babesiosis. These diseases are of increasing public
 health concern, especially as the geographic range and seasonality of
@@
 
-Looked for Missing Data & General Trends
+**Looked for Missing Data & General Trends**
 
--   Identified Disease Burden Trends & Hotspot Counties
+-   Excluded cases with a missing or non-“CONFIRMED” STATUS to retain
+    only confirmed cases.
+-   Refined the dataset by retaining only confirmed “positive” test
+    results, excluding missing or ambiguous results, leaving 95 valid
+    cases.
+-   Additional filtering ensured correct identification of positive
+    cases while excluding non-positive values.
 
-<!-- -->
+**Identified Disease Burden Trends & Hotspot Counties**
 
--   Generated a stacked bar graph to visualize case distribution across
-    regions. - Checked Age Distribution
+-   Generated a histogram to visualize the distribution of tick-borne
+    diseases in Georgia between 2019 and 2024.
+-   Created a bar chart highlighting the five most common diseases.
+-   Specifically analyzed Lyme Disease cases, focusing on test results.
 
-<!-- -->
+**Checked Age Distribution**
 
@@
     dataset.
-
-<!-- -->
-
 -   Conducted an exploratory analysis of disease counts by age group.
 
--   Evaluated Data Completeness Over Time
+**Evaluated Data Completeness Over Time**
 
 -   Extracted onset date information to track missing data trends.
+-   Created a plot distinguishing between blank and non-blank STATUS
+    entries over time to identify periods of high missingness,
+    indicating possible data entry errors or reporting gaps.
 
--   Created a plot distinguishing between blank and non-blank STATUS
-    entries over time.
-
--   Identified periods of high missingness, which may indicate data
-    entry errors or reporting gaps.
-
-Analyzed Timeline Between Key Case Events
+**Analyzed Timeline Between Key Case Events**
 
 -   Calculated time differences for:
+    -   Onset date (DOO) to first lab test date (LABDATE1)
+    -   First lab test (LABDATE1) to second lab test (LABDATE2)
+    -   Second lab test (LABDATE2) to case update date (UPDATEDATE)
+-   Summed these values to determine the total case duration.
+-   Saved the processed dataset as a CSV file for further analysis.
+-   Created a formatted HTML table to present the extracted time
+    intervals.
 
-    -   Onset date (DOO) to first lab test date (LABDATE1)
+**Addressed Data Quality Issues**
 
-    -   First lab test (LABDATE1) to second lab test (LABDATE2)
-
-    -   Second lab test (LABDATE2) to case update date (UPDATEDATE)
-
-    -   Summed these values to determine the total case duration. -
-        Saved the processed dataset as a CSV file for further
-        analysis. - Created a formatted HTML table to present extracted
-        time intervals.
-
-Addressed Data Quality Issues
-
--   Identified negative time values in lab test dates, indicating data
-    entry errors
-
+-   Identified negative time values between lab test dates, indicating
+    data entry errors.
 -   Implemented a filtering step to remove records with invalid date
-    sequences
+    sequences.
 
@@
 
+# Results
+
+## Exploratory/Descriptive analysis
+
+*Use a combination of text/tables/figures to explore and describe your
+data. Show the most important descriptive results here. Additional ones
+should go in the supplement. Even more can be in the R and Quarto files
+that are part of your project.*
+
+Note the loading of the data providing a **relative** path using the
+`../../` notation. (Two dots means a folder up). You never want to
+specify an **absolute** path like `C:\ahandel\myproject\results\`
+because if you share this with someone, it won't work for them since
+they don't have that path. You can also use the `here` R package to
+create paths. See examples of that below. I generally recommend the
+`here` package.
+
 **Rocky Mountain Spotted Fever (RMSF) Severity:**
@@
 
-Given its similarity to Alpha-Gal Syndrome
-(which is of interest to me and the topic of my capstone), Lyme Disease
-cases are analyzed separately. A count histogram compares positive,
-negative, and other test outcomes, providing insights into testing
-patterns and diagnosis rates.
+Given its similarity to Alpha-Gal Syndrome (which is of interest to me
+and the topic of my capstone), Lyme Disease cases are analyzed
+separately. A count histogram compares positive, negative, and other
+test outcomes, providing insights into testing patterns and diagnosis
+rates.
 
@@
 
-A grouped bar chart maps case counts across
-Georgia counties, categorized by disease type. Due to the density of
-information, this visualization is refined to focus on the top five
-tick-borne diseases and their distribution across the 12 counties with
-the highest confirmed case counts.
+A grouped bar chart maps case counts across Georgia counties,
+categorized by disease type. Due to the density of information, this
+visualization is refined to focus on the top five tick-borne diseases
+and their distribution across the 12 counties with the highest confirmed
+case counts.
+
 ```{r}
@@
 
-Age distribution is assessed as part of
-the initial data cleaning process, confirming that no cases in the
-cleaned dataset have missing age values. Additionally, an exploratory
-analysis examines disease counts by age group to identify potential
-patterns in disease prevalence.
+Age distribution is assessed as part of the initial data cleaning
+process, confirming that no cases in the cleaned dataset have missing
+age values. Additionally, an exploratory analysis examines disease
+counts by age group to identify potential patterns in disease
+prevalence.
+
 ```{r}
@@
 **Data Completeness Over Time**
-To evaluate how data completeness has
-evolved, the dataset is processed to extract onset date information. A
-visualization distinguishes between blank and non-blank entries in the
-STATUS column over time, helping to identify periods with high
-missingness. Spikes in blank values may indicate data entry errors,
-system issues, or reporting gaps that should be addressed before
-conducting further analysis.
+
+To evaluate how data completeness has evolved, the dataset is processed
+to extract onset date information. A visualization distinguishes between
+blank and non-blank entries in the STATUS column over time, helping to
+identify periods with high missingness. Spikes in blank values may
+indicate data entry errors, system issues, or reporting gaps that should
+be addressed before conducting further analysis.
 
@@
 
-**Onset date (DOO) to first lab test date (LABDATE1)**
+-   **Onset date (DOO) to first lab test date (LABDATE1)**
 
-**First lab test (LABDATE1) to second lab test (LABDATE2)**
+-   **First lab test (LABDATE1) to second lab test (LABDATE2)**
 
-**Second lab test (LABDATE2) to case update date (UPDATEDATE)**
+-   **Second lab test (LABDATE2) to case update date (UPDATEDATE)**
 
@@
 
-*Explain anything related to your statistical analyses.*
-
 {{< pagebreak >}}
 
-# Results
+## Basic statistical analysis
 
-## Exploratory/Descriptive analysis
+For the first key question related to the disease trends over time (in
+years). In looking at the plot, it appears that there have been
+increases over time were observed in most of the diseases. However, the
+rate of increase was relatively slow as indicated by the lack of
+statistical significance of these trends (p\>0.5) which may indicate an
+attribution to the increase in reporting rather than an increase in
+overall disease burden. In the full analysis and second iteration of
+this, these trends will be compared year to year, comparing each year to
+the 2018 standard.
 
-*Use a combination of text/tables/figures to explore and describe your
-data. Show the most important descriptive results here. Additional ones
-should go in the supplement. Even more can be in the R and Quarto files
-that are part of your project.*
++------------+------------+------------+------------+------------+
+| term       | estimate   | std.error  | statistic  | p.value    |
+|            |            |            |            |            |
+| \<chr\>    | \<dbl\>    | \<dbl\>    | \<dbl\>    | \<dbl\>    |
++:===========+===========:+===========:+===========:+===========:+
+| (          | -2         | 1          | -          | 0.1729390  |
+| Intercept) | 3          | 573.052665 | 1.48025251 |            |
+|            | 28.5151515 |            |            |            |
++------------+------------+------------+------------+------------+
+| YEAR       | 1.1515152  | 0.777581   | 1.48089418 | 0.1727716  |
++------------+------------+------------+------------+------------+
+| DISEAS     | 0.4242424  | 5.484565   | 0.07735207 | 0.9400358  |
+| E          |            |            |            |            |
+| BABESIOSIS |            |            |            |            |
++------------+------------+------------+------------+------------+
+| DI         | 4.8484848  | 6.364775   | 0.76176850 | 0.4656907  |
+| S          |            |            |            |            |
+| EASEDENGUE |            |            |            |            |
++------------+------------+------------+------------+------------+
+| DISEA      | 5.0000000  | 5.157889   | 0.96938889 | 0.3576794  |
+| S          |            |            |            |            |
+| EEHRLICHIA |            |            |            |            |
+| C          |            |            |            |            |
+| HAFFEENSIS |            |            |            |            |
++------------+------------+------------+------------+------------+
+| DISEA      | -1.1515152 | 6.364775   | -          | 0.8604390  |
+| S          |            |            | 0.18092002 |            |
+| EEHRLICHIA |            |            |            |            |
+| EWINGII    |            |            |            |            |
++------------+------------+------------+------------+------------+
+| D          | 8.5606061  | 4.963745   | 1.72462659 | 0.1186814  |
+| ISEASELYME |            |            |            |            |
+| DISEASE    |            |            |            |            |
++------------+------------+------------+------------+------------+
+| D          | 0.5757576  | 5.484565   | 0.10497781 | 0.9186958  |
+| I          |            |            |            |            |
+| SEASEROCKY |            |            |            |            |
+| MOUNTAIN   |            |            |            |            |
+| SPOTTED    |            |            |            |            |
+| FEVER      |            |            |            |            |
++------------+------------+------------+------------+------------+
+| D          | 8.2272727  | 5.593721   | 1.47080497 | 0.1754208  |
+| ISEASEWEST |            |            |            |            |
+| NILE (WNV) |            |            |            |            |
+| INFECTION  |            |            |            |            |
++------------+------------+------------+------------+------------+
 
-Note the loading of the data providing a **relative** path using the
-`../../` notation. (Two dots means a folder up). You never want to
-specify an **absolute** path like `C:\ahandel\myproject\results\`
-because if you share this with someone, it won't work for them since
-they don't have that path. You can also use the `here` R package to
-create paths. See examples of that below. I generally recommend the
-`here` package.
+9 rows
 
-## Basic statistical analysis
+```{r}
+library(here)
+library(knitr)
+
+# Generate correct file path
+img_path <- here::here("results", "figures", "trend_over_time_by_disease.png")
+
+# Display the image
+knitr::include_graphics(img_path)
+```
+
+The most interesting finding from the initial modeling was in relation
+to the data completeness research question. In modeling the differences
+across time in whether a final STATUS was inputted rather than left
+blank, a general decrease was seen across the full 2018-2024 time scale.
+Specifically, completion rates of the reporting form were approximately
+50% in 2018 but dropped rapidly to 20% in 2019-- around which the rate
+of completeness has hovered since 2019. However, in linear modeling, the
+changes across time in the aggregate, p \> 0.05.
+
+```{r}
+library(here)
+library(knitr)
+
+# Generate correct file path
+img_path <- here::here("results", "figures", "completeness_over_time.png")
+
+# Display the image
+knitr::include_graphics(img_path)
+
+```
+
++------------+------------+------------+------------+------------+
+| term       | estimate   | std.error  | statistic  | p.value    |
+|            |            |            |            |            |
+| \<chr\>    | \<dbl\>    | \<dbl\>    | \<dbl\>    | \<dbl\>    |
++:===========+===========:+===========:+===========:+===========:+
+| (          | 9          | 5          | 1.733920   | 0.1434648  |
+| Intercept) | 935.267857 | 729.946482 |            |            |
++------------+------------+------------+------------+------------+
+| YEAR       | -4.910714  | 2.835202   | -1.732051  | 0.1438108  |
++------------+------------+------------+------------+------------+
+
+2 rows
+
+On a positive note for the utilization of the reporting system and
+trends towards better tick-bourne disease response, there has been
+statistically significant decreases in the time between oneset and an
+initial lab visit/interaction. In 2018, the average time between these
+two occurrences ranged from approximately 150-520 days. With marked
+decreases, the 2024 data indicates that patients with tick-bourne
+disease occurrences may be interacting with providers or lab testing in
+the same month, same week or even same day.
+
+```{r}
+library(here)
+library(knitr)
+
+# Generate correct file path
+img_path <- here::here("results", "figures", "days_to_lab_over_time.png")
+
+# Display the image
+knitr::include_graphics(img_path)
+```
 

### 2025-04-21 code/analysis-code/statistical-analysis-gini-index.qmd
Part 4/5
Catching Up - Part Four & Part Five, citations still need to be figured out and some more fun graphics (my findings are boring) but closer?

---
title: "Data Analysis for Raw Tick-Bourne Cases Data - 2019 to 2024"
author: "Hope Grismer"
date: "2025-03-10"
output: html_document
---

```{r}
## ---- setup --------
library(tidymodels)
library(readr)
library(dplyr)
library(here)

## ---- load_data --------
# Load cleaned modeling dataset
model_data <- read_csv(here("data", "processed-data", "time_between_processeddata.csv"))

# Load Gini index data (must have columns COUNTY and GINI_INDEX)
gini_data <- read_csv(here("data", "external-data", "gini_index_by_county.csv"))

# Merge datasets on COUNTY
model_data <- model_data %>%
  left_join(gini_data, by = "COUNTY") %>%
  mutate(
    delay_over_threshold = as.factor(delay_over_threshold),
    COUNTY = as.factor(COUNTY)
  ) %>%
  drop_na(delay_over_threshold, COUNTY, GINI_INDEX)

## ---- data_split --------
set.seed(123)
data_split <- initial_split(model_data, prop = 0.8, strata = delay_over_threshold)
train_data <- training(data_split)
test_data <- testing(data_split)

## ---- recipe --------
delay_recipe <- recipe(delay_over_threshold ~ COUNTY + GINI_INDEX, data = train_data) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_normalize(all_numeric_predictors())

## ---- models --------
logistic_model <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")

tree_model <- decision_tree() %>%
  set_engine("rpart") %>%
  set_mode("classification")

rf_model <- rand_forest(trees = 500) %>%
  set_engine("ranger") %>%
  set_mode("classification")

## ---- workflows --------
logistic_workflow <- workflow() %>%
  add_recipe(delay_recipe) %>%
  add_model(logistic_model)

tree_workflow <- workflow() %>%
  add_recipe(delay_recipe) %>%
  add_model(tree_model)

rf_workflow <- workflow() %>%
  add_recipe(delay_recipe) %>%
  add_model(rf_model)

## ---- cross_validation --------
set.seed(123)
cv_folds <- vfold_cv(train_data, v = 5, repeats = 3, strata = delay_over_threshold)

# Fit all models
logistic_res <- fit_resamples(logistic_workflow, resamples = cv_folds, metrics = metric_set(accuracy, roc_auc))
tree_res     <- fit_resamples(tree_workflow, resamples = cv_folds, metrics = metric_set(accuracy, roc_auc))
rf_res       <- fit_resamples(rf_workflow, resamples = cv_folds, metrics = metric_set(accuracy, roc_auc))

## ---- compare_models --------
collect_metrics(logistic_res)
collect_metrics(tree_res)
collect_metrics(rf_res)

## ---- final_evaluation --------
final_rf_fit <- rf_workflow %>%
  last_fit(data_split)

# Metrics
collect_metrics(final_rf_fit)

# Confusion matrix
conf_mat <- collect_predictions(final_rf_fit) %>%
  conf_mat(truth = delay_over_threshold, estimate = .pred_class)

# ROC plot
roc_plot <- collect_predictions(final_rf_fit) %>%
  roc_curve(truth = delay_over_threshold, .pred_1) %>%
  autoplot() +
  theme_minimal() +
  labs(title = "ROC Curve – Random Forest (County + Gini Only)")

ggsave(
  filename = here("results", "figures", "roc_curve_gini_model.png"),
  plot = roc_plot,
  width = 6, height = 5, dpi = 300
)

# Save confusion matrix plot (optional)
conf_plot <- autoplot(conf_mat, type = "heatmap") +
  scale_fill_gradient(low = "white", high = "#2C3E50") +
  theme_minimal() +
  labs(title = "Confusion Matrix – Random Forest")

ggsave(
  filename = here("results", "figures", "conf_matrix_gini_model.png"),
  plot = conf_plot,
  width = 5, height = 5, dpi = 300
)

```

### 2025-04-21 code/analysis-code/statistical-analysis-poverty-rate.qmd
Part 4/5
Catching Up - Part Four & Part Five, citations still need to be figured out and some more fun graphics (my findings are boring) but closer?

```{r}
library(tidyverse)
library(tidymodels)
library(here)

# STEP 1: Summarize delay rate per county
delay_summary <- read_csv(here("data", "processed-data", "time_between_processeddata.csv"), show_col_types = FALSE) %>%
  mutate(COUNTY = str_to_title(COUNTY)) %>%
  group_by(COUNTY) %>%
  summarise(
    total_cases = n(),
    delay_cases = sum(TOTAL_DAYS > 30, na.rm = TRUE),
    delay_rate = delay_cases / total_cases
  ) %>%
  mutate(delay_rate_over_30 = as.factor(if_else(delay_rate > 0.3, 1, 0)))  # Binary outcome

# STEP 2: Load and prepare poverty rate data
poverty_data <- read_csv(here("data", "raw-data", "poverty_rate_by_county.csv"), show_col_types = FALSE) %>%
  transmute(
    COUNTY = str_remove(str_to_title(County), " County$"),
    POVERTY_RATE = as.numeric(POVERTY_RATE)
  )

# STEP 3: Merge datasets and clean
model_data <- delay_summary %>%
  left_join(poverty_data, by = "COUNTY") %>%
  drop_na(delay_rate_over_30, POVERTY_RATE)

# STEP 4: Define preprocessing recipe
model_recipe <- recipe(delay_rate_over_30 ~ POVERTY_RATE, data = model_data) %>%
  step_normalize(all_numeric_predictors())

# STEP 5: Define models
logistic_model <- logistic_reg() %>%
  set_engine("glm") %>%
  set_mode("classification")

rf_model <- rand_forest(trees = 500) %>%
  set_engine("ranger") %>%
  set_mode("classification")

# STEP 6: Create workflows
logistic_workflow <- workflow() %>%
  add_recipe(model_recipe) %>%
  add_model(logistic_model)

rf_workflow <- workflow() %>%
  add_recipe(model_recipe) %>%
  add_model(rf_model)

# STEP 7: Cross-validation setup
set.seed(123)
cv_folds <- vfold_cv(model_data, v = 5, repeats = 3, strata = delay_rate_over_30)

# STEP 8: Fit models
logistic_res <- fit_resamples(logistic_workflow, resamples = cv_folds, metrics = metric_set(accuracy, roc_auc))
rf_res <- fit_resamples(rf_workflow, resamples = cv_folds, metrics = metric_set(accuracy, roc_auc))

# STEP 9: Collect performance metrics
logistic_metrics <- collect_metrics(logistic_res)
rf_metrics <- collect_metrics(rf_res)

# Display metrics
logistic_metrics
rf_metrics
```

### 2025-04-21 code/analysis-code/statistical-analysis.qmd
Part 4/5
Catching Up - Part Four & Part Five, citations still need to be figured out and some more fun graphics (my findings are boring) but closer?

@@
 # Save plot
-plot_file1 <- here::here("results", "figures", "trend_over_time_by_disease.png")
+plot_file1 <- here::here("results", "supplemental-figures", "trend_over_time_by_disease.png")
 ggsave(plot_file1, plot = plot1, width = 8, height = 6, dpi = 300)
@@
 ```{r}
-# 2. Distribution Across Health Districts:
-# Bar plot of cases by health district
+##Distribution Across Counties:
+# Bar plot of cases by county
 plot2 <- ggplot(cleaned_data3, aes(x = COUNTY, fill = DISEASE)) +
@@
   labs(
-    title = "Cases by Health District",
-    x = "Health District",
+    title = "Cases by County",
+    x = "County",
     y = "Number of Cases"
@@
 # Save plot
-plot_file2 <- here::here("results", "figures", "cases_by_health_district.png")
+plot_file2 <- here::here("results", "supplemental-figures", "cases_by_eachcounty.png")
 ggsave(plot_file2, plot = plot2, width = 8, height = 6, dpi = 300)
@@
 
-# Save the plot to the results/figures folder
-plot_file3 <- here::here("results", "figures", "completeness_over_time.png")
+# Save the plot 
+plot_file3 <- here::here("results", "supplemental-figures", "completeness_over_time.png")
 ggsave(plot_file3, plot = plot3, width = 8, height = 6, dpi = 300)
@@
 # Save plot
-plot_file4 <- here::here("results", "figures", "days_to_lab_over_time.png")
+plot_file4 <- here::here("results", "supplemental-figures", "days_to_lab_over_time.png")
 ggsave(plot_file4, plot = plot4, width = 8, height = 6, dpi = 300)
@@
 # Save plot
-plot_file5 <- here::here("results", "figures", "relative_incidence_2019_2024.png")
+plot_file5 <- here::here("results", "supplemental-figures", "relative_incidence_2019_2024.png")
 ggsave(plot_file5, plot = plot5, width = 8, height = 6, dpi = 300)
 ```
+
+```{r}
+## ---- modeling_setup --------
+library(tidymodels)
+
+# Load the processed dataset
+model_data <- readr::read_csv(here::here("data", "processed-data", "time_between_processeddata.csv"))
+
+# Add the target variable and convert appropriate columns
+model_data <- model_data %>%
+  mutate(
+    delay_over_threshold = if_else(TOTAL_DAYS > 30, 1, 0),  # Create binary target
+    delay_over_threshold = as.factor(delay_over_threshold),  # Convert to factor
+    DISEASE = as.factor(DISEASE),
+    COUNTY = as.factor(COUNTY),
+    SEX = as.factor(GENDER),
+    AGE = as.numeric(AGE)  # Ensure numeric
+  ) %>%
+  drop_na(delay_over_threshold, AGE, SEX, COUNTY, DISEASE)  # Clean up
+
+# Save the updated dataset with the new variable
+write_csv(model_data, here::here("data", "processed-data", "time_between_processeddata.csv"))
+
+```
+
+To explore whether certain demographic or clinical variables could predict delayed follow-up in confirmed tick-borne disease cases, we created a binary outcome variable called delay_over_threshold, coded as 1 if the total time between disease onset and case closure exceeded 30 days, and 0 otherwise.
+This variable was calculated based on the sum of time intervals between key date fields (DOO, LABDATE1, LABDATE2, and UPDATEDATE), which had been previously cleaned and processed.
+Predictors selected for this analysis included age, sex, county, and disease type, as these are commonly available surveillance variables that may relate to access to care or follow-up procedures.
+All categorical predictors were converted into factor variables to prepare them for modeling.
+
+```{r}
+## ---- modeling_setup --------
+library(tidymodels)
+
+# Load the processed dataset
+model_data <- readr::read_csv(here::here("data", "processed-data", "time_between_processeddata.csv"))
+
+# Convert appropriate variables to factors
+model_data <- model_data %>%
+  mutate(
+    delay_over_threshold = as.factor(delay_over_threshold),  # target variable
+    DISEASE = as.factor(DISEASE),
+    COUNTY = as.factor(COUNTY),
+    SEX = as.factor(GENDER),
+    AGE = as.numeric(AGE)  # just in case it's not numeric yet
+  ) %>%
+  drop_na(delay_over_threshold, AGE, SEX, COUNTY, DISEASE)  # Remove any rows with missing predictors
+
+```
+
+The data was randomly partitioned into training and testing sets using an 80/20 split, stratified by the outcome variable to maintain a balanced class distribution.
+The training set was used for model training and cross-validation, while the test set was reserved for final evaluation of model performance on unseen data.
+
+```{r}
+set.seed(123)
+data_split <- initial_split(model_data, prop = 0.8, strata = delay_over_threshold)
+train_data <- training(data_split)
+test_data <- testing(data_split)
+
+```
+
+A recipe() was used to define preprocessing steps applied uniformly across all models.
+Categorical variables (such as county, disease, and sex) were one-hot encoded using dummy variables.
+Zero-variance predictors were removed, and the single numeric predictor, age, was normalized to ensure comparability across different model types.
+```{r}
+delay_recipe <- recipe(delay_over_threshold ~ AGE + SEX + COUNTY + DISEASE, data = train_data) %>%
+  step_dummy(all_nominal_predictors()) %>%  # convert categorical predictors to dummy variables
+  step_zv(all_predictors()) %>%  # remove zero-variance predictors
+  step_normalize(all_numeric_predictors())  # standardize age
+```
+
+```{r}
+logistic_model <- logistic_reg() %>%
+  set_engine("glm") %>%
+  set_mode("classification")
+
+tree_model <- decision_tree() %>%
+  set_engine("rpart") %>%
+  set_mode("classification")
+
+rf_model <- rand_forest(trees = 500) %>%
+  set_engine("ranger") %>%
+  set_mode("classification")
+
+```
+
+For each model, a workflow was constructed using the tidymodels framework to combine the preprocessing recipe and the corresponding model specification.
+This approach ensured consistent preprocessing and simplified model training and evaluation.
+```{r}
+logistic_workflow <- workflow() %>%
+  add_model(logistic_model) %>%
+  add_recipe(delay_recipe)
+
+tree_workflow <- workflow() %>%
+  add_model(tree_model) %>%
+  add_recipe(delay_recipe)
+
+rf_workflow <- workflow() %>%
+  add_model(rf_model) %>%
+  add_recipe(delay_recipe)
+```
+To evaluate model performance and avoid overfitting, we implemented 5-fold cross-validation repeated 3 times.
+This resampling strategy provides a robust estimate of out-of-sample performance.
+Each model was trained and evaluated on multiple folds using two performance metrics: accuracy and the area under the ROC curve (AUC), the latter of which is especially informative for imbalanced or binary classification problems.
+```{r}
+set.seed(123)
+cv_folds <- vfold_cv(train_data, v = 5, repeats = 3, strata = delay_over_threshold)
+
+# Fit all models
+logistic_res <- fit_resamples(logistic_workflow, resamples = cv_folds, metrics = metric_set(roc_auc, accuracy))
+tree_res     <- fit_resamples(tree_workflow, resamples = cv_folds, metrics = metric_set(roc_auc, accuracy))
+rf_res       <- fit_resamples(rf_workflow, resamples = cv_folds, metrics = metric_set(roc_auc, accuracy))
+
+```
+Cross-validation results for each model were collected and compared using mean accuracy and AUC.
+These metrics provided insight into the predictive quality of each approach.
+The model with the best combination of high accuracy and AUC was selected for final evaluation on the test set.
+```{r}
+collect_metrics(logistic_res)
+collect_metrics(tree_res)
+collect_metrics(rf_res)
+```
+
+```{r}
+## ---- final_rf_evaluation --------
+# Fit the final random forest model on training data and evaluate on test data
+final_rf_fit <- rf_workflow %>%
+  last_fit(split = data_split)
+
+# Collect performance metrics
+collect_metrics(final_rf_fit)
+
+# Confusion matrix
+collect_predictions(final_rf_fit) %>%
+  conf_mat(truth = delay_over_threshold, estimate = .pred_class)
+
+# ROC curve
+collect_predictions(final_rf_fit) %>%
+  roc_curve(truth = delay_over_threshold, .pred_1) %>%
+  autoplot()
+
+# Save ROC curve plot
+roc_plot <- collect_predictions(final_rf_fit) %>%
+  roc_curve(truth = delay_over_threshold, .pred_1) %>%
+  autoplot() +
+  theme_minimal() +
+  labs(title = "ROC Curve for Random Forest Model")
+
+ggsave(
+  filename = here::here("results", "supplemental-figures", "roc_curve.png"),
+  plot = roc_plot,
+  width = 6, height = 5, dpi = 300
+)
+```

### 2025-04-21 code/processing-code/processing-code-tickcases.qmd
Part 4/5
Catching Up - Part Four & Part Five, citations still need to be figured out and some more fun graphics (my findings are boring) but closer?

@@
 # Create histogram of disease counts
-disease_plot <- ggplot(cleaned_data3, aes(x = DISEASE)) +
+disease_histogram <- ggplot(cleaned_data3, aes(x = DISEASE)) +
   geom_bar(fill = "#2C3E50", color = "white") +  # Dark blue bars with white edges
@@
 # Define save locations
-save_plot_location <- here::here("results", "figures", "disease_histogram.png")
+save_plot_location <- here::here("results", "supplemental-figures", "disease_histogram.png")
 
 # Ensure directory exists
-dir.create(here::here("results", "figures"), recursive = TRUE, showWarnings = FALSE)
+dir.create(here::here("results", "supplemental-figures"), recursive = TRUE, showWarnings = FALSE)
 
 # Save the plot
-ggsave(filename = save_plot_location, plot = disease_plot, width = 12, height = 6, dpi = 300)
+ggsave(filename = save_plot_location, plot = disease_histogram, width = 12, height = 6, dpi = 300)
 ```
@@
 # Define save location
-save_rmsf_plot_location <- here::here("results", "figures", "rmsf_titer_boxplot_log.png")
+save_rmsf_plot_location <- here::here("results", "supplemental-figures", "rmsf_titer_boxplot_log.png")
 
@@
 # Define save location
-save_top5_disease_plot <- here::here("results", "figures", "top5_diseases_plot.png")
+save_top5_disease_plot <- here::here("results", "supplemental-figures", "top5_diseases_plot.png")
 
@@
 # Define save location
-save_lyme_count_plot <- here::here("results", "figures", "lyme_result_count_histogram2.png")
+save_lyme_count_plot <- here::here("results", "supplemental-figures", "lyme_result_count_histogram2.png")
 
@@
 # Define save location
-save_disease_county_plot <- here::here("results", "figures", "disease_county_plot.png")
+save_disease_county_plot <- here::here("results", "supplemental-figures", "disease_county_plot.png")
 
@@
 # Define save location
-save_top5_stacked_plot <- here::here("results", "figures", "top5_diseases_stacked_plot.png")
+save_top5_stacked_plot <- here::here("results", "supplemental-figures", "top5_diseases_stacked_plot.png")
 
@@
 
-# Save the plot to results/figures using here::here()
+# Save the plot to results/supplemental-figures using here::here()
 ggsave(
-  filename = here::here("results", "figures", "disease_count_by_age.png"), 
+  filename = here::here("results", "supplemental-figures", "disease_count_by_age.png"), 
   plot = age_group_plot, 
@@
 
-# Save the plot to results/figures using here::here()
+# Save the plot to results/supplemental-figures using here::here()
 ggsave(
-  filename = here::here("results", "figures", "status_completeness_plot.png"), 
+  filename = here::here("results", "supplemental-figures", "status_completeness_plot.png"), 
   plot = status_completeness_plot, 
@@
 
+# Create binary target variable for modeling: delay > 30 days = 1, else 0
+processed_time_data2 <- processed_time_data2 %>%
+  mutate(
+    delay1 = as.numeric(LABDATE1 - DOO),
+    delay2 = as.numeric(LABDATE2 - LABDATE1),
+    delay3 = as.numeric(UPDATEDATE - LABDATE2),
+    total_delay_days = delay1 + delay2 + delay3,
+    delay_over_threshold = if_else(total_delay_days > 30, 1, 0)
+  )
+
 # Select key columns for review

### 2025-04-21 products/manuscript/Manuscript.qmd
Part 4/5
Catching Up - Part Four & Part Five, citations still need to be figured out and some more fun graphics (my findings are boring) but closer?

--- 2025-03-02 products/manuscript/Manuscript.qmd
@@
 
-The structure below is one possible setup for a manuscript, or a general
-data analysis project (including the course project). Adjust as needed.
-You don't need to have exactly these sections, but the content covering
-those sections should be addressed.
-
-This uses MS Word as output format. [See
-here](https://quarto.org/docs/output-formats/ms-word.html) for more
-information. You can switch to other formats, like html or pdf. See [the
-Quarto documentation](https://quarto.org/) for other formats.
-
 ```{r, echo=FALSE, message=FALSE}
@@
 
-**Authors**
+Authors
 
 -   Hope Grismer $^{1,2,*, \land}$ (ORCID: 0000-0000-1234-5678)
--   Second Author$^{3, *}$
--   Third Author$^{3, \dagger}$
--   Last Author$^{1, 2,}$
 
-**Author affiliations**
+Author affiliations
 
@@
     (SECVBD), Gainesville, FL 32611.
-3.  Yet another Affiliation.
-
-$*$ These authors contributed equally to this work.
 
@@
 
-# Summary/Abstract
+# Abstract
 
-*Write a summary of your project.*
+This project explores patterns in tick-borne disease cases reported to
+the Georgia Department of Public Health (DPH) between 2019 and 2024. The
+analysis involved extensive data cleaning to retain only confirmed and
+positive cases, with additional processing to resolve inconsistencies in
+timelines and reporting. Exploratory data analysis (EDA) was used to
+examine disease distribution, demographic patterns, geographic hotspots,
+and data completeness over time. Building on these insights, predictive
+models were developed to assess whether patient and case-level
+factors—such as age, disease type, and county—could help identify cases
+at risk of delayed follow-up. This work theorectically supports improved
+understanding of disease surveillance data and highlights areas for
+strengthening tick-borne disease reporting and case management systems
+in Georgia.
 
@@
 
-## General Background Information
+### 2.1 Background
+
+In the past two decades, the study of tick-borne disease has gained
+increasing attention due to rising incidence rates and expanding
+geographic ranges of disease-carrying ticks. While early medical and
+veterinary research in the 20th century identified key tick-borne
+illnesses such as Rocky Mountain spotted fever and Lyme disease, recent
+decades have seen an acceleration in the detection of new pathogens,
+greater public health concern, and expanded surveillance efforts across
+the globe. Researchers have documented the ecology and behavior of
+ticks, explored the role of climate and land use in tick habitat
+expansion, and analyzed patterns of disease incidence in both human and
+animal populations. Despite the growing body of research, many knowledge
+gaps remain, particularly in understanding co-infections, long-term
+health effects, and effective prevention strategies. Tick-borne diseases
+(TBDs) are primarily transmitted through the bite of infected
+hard-bodied ticks (family Ixodidae), and include bacterial, viral, and
+protozoan pathogens. Today, TBDs are recognized as a major public health
+concern in both rural and suburban settings, with transmission risk
+influenced by ecological, climatic, and sociocultural factors.
+
+### 2.1.1 General Background Information
 
@@
 or parasites—that can then be transmitted to humans. Examples of
-tick-bourne diseases include, but are not limited to, Lyme disease,
-anaplasmosis, and babesiosis. These diseases are of increasing public
+tick-borne diseases include, but are not limited to, Lyme disease,
+Anaplasmosis, and Babesiosis. These diseases are of increasing public
 health concern, especially as the geographic range and seasonality of
@@
 
-## Description of data and data source
+### 2.1.2 Transmission and Reservoirs
+
+Ticks act as both vectors and reservoirs for a variety of pathogens,
+transmitting disease to humans and animals during blood meals. Most
+tick-borne diseases are zoonotic, with small mammals, deer, and birds
+acting as key reservoirs in natural cycles. Transmission to humans
+typically occurs when a nymph or adult tick feeds for an extended
+period, allowing the pathogen to transfer from the tick’s salivary
+glands into the host’s bloodstream. Common tick-borne pathogens include
+Borrelia burgdorferi (the agent of Lyme disease), Rickettsia rickettsii
+(Rocky Mountain spotted fever), Anaplasma phagocytophilum, Ehrlichia
+chaffeensis, and viruses such as Powassan and Heartland virus. Ticks
+themselves are influenced by multiple environmental factors, including
+temperature, humidity, vegetation, and host availability. Human exposure
+risk is increased in areas with high host density, edge habitats between
+forest and residential zones, and during seasons with peak tick
+activity. Additionally, land use changes—such as suburban development,
+deforestation, and agricultural expansion—can modify tick habitats and
+affect pathogen dynamics.
+
+### 2.1.3 Ecological and Health Impacts
+
+The ecological consequences of tick-borne disease extend beyond human
+health. Tick populations and their associated pathogens can alter
+wildlife health and community composition. Infected animals may suffer
+sublethal effects that reduce fitness, while high burdens of tick
+parasitism can influence population dynamics, particularly in rodents
+and ungulates. From a public health perspective, TBDs can cause a range
+of acute and chronic symptoms in humans, from fever and fatigue to
+long-term neurological and musculoskeletal complications.
+Co-infections—when an individual is infected with more than one
+pathogen—are increasingly recognized and may complicate diagnosis and
+treatment. The public health burden is also shaped by disparities in
+access to health care, awareness of TBDs, and availability of diagnostic
+testing, which may result in underreporting or misdiagnosis. The costs
+associated with tick-borne disease—including medical care, productivity
+loss, and prevention efforts—represent a growing concern for health
+systems, particularly in endemic areas.
+
+### 2.1.4 Relevance
+
+Tick-borne diseases have been documented in all regions of the United
+States, with the southeastern, northeastern, and upper midwestern states
+experiencing particularly high incidence rates. Climate models suggest
+that warming temperatures and shifting precipitation patterns may expand
+the range of ticks and lengthen their active season, potentially
+increasing the geographic footprint and seasonality of tick-borne
+disease transmission. In the southeastern U.S., for example, Amblyomma
+americanum (the lone star tick) has expanded in abundance and geographic
+range over the past decades, contributing to increased cases of
+ehrlichiosis and the emergence of Alpha-Gal Syndrome, a red meat allergy
+triggered by tick bites. Understanding the ecology, distribution, and
+drivers of tick-borne disease is therefore crucial to predicting future
+trends, identifying at-risk populations, and developing effective
+prevention and response strategies.
+
+# 3. The Data
+
+# 3.1 Data Acquisition & Description
+
+The data used in this analysis was obtained through my internship with
+the Georgia Department of Health, in collaboration with the Centers for
+Disease Control and Prevention (CDC). Specifically, it is derived from
+the Nationally Notifiable Diseases Surveillance System (NNDSS) and
+focuses on tick-borne illnesses. This data set provides key insights
+into the trends, distribution, and characteristics of reported cases,
+supporting public health efforts in disease monitoring and
+prevention.2.1.5 Description of Data and Data Source
 
@@
 Department of Public Health (GA-DPH) between 2019 and 2024. It was
-graciously provided by DPH professionals in the epidemiology department.
-All available data for each case over this five-year period across the
-state of Georgia. Proper de-identification practices-- the removal of
-first name, last name, and date of birth (DOB)-- has been taken. The
-dataset includes variables such as disease type, patient demographics
-(gender, age, district, state, county, and ZIP code), medical details
-(incident ID, report ID, lab test dates, results, serotype, and titers),
-as well as disease outcomes (death status, fever, migranes, and
-administrative status). Reported diagnoses included in this dataset
-include Toxoplasmosis, Dengue, Rocky Mountain Spotted Fever, Anaplasma
-Phagocytophilum, Malaria, Lyme Disease, West Nile (WNV) Infection,
-Chikungunya, LaCrosse (LAC) Infection, Zika, Ehrlichia Chaffeensis, and
-Typhus (Epidemic/Louse/Flea). Some of this cases, based on adherence to
-specific case definitions and subsequent lab results, are categorized as
-suspect, probable, confirmed, or unconfirmed. By examining case reports,
-public health professionals can better understand the prevalence,
-demographics, and regional variations of these diseases. This analysis
-aims to further inform the public health understanding of tick-bourne
-disease in Georgia.
+graciously provided by DPH professionals in the epidemiology department
+and includes all available data for each case reported across the state
+over this five-year period. Proper de-identification practices—such as
+removal of first name, last name, and date of birth (DOB)—were followed.
+The dataset includes variables such as disease type, patient
+demographics (gender, age, district, state, county, and ZIP code),
+medical details (incident ID, report ID, lab test dates, results,
+serotype, and titers), and disease outcomes (death status, fever,
+migraines, and administrative status). Reported diagnoses include
+Toxoplasmosis, Dengue, Rocky Mountain Spotted Fever, Anaplasma
+phagocytophilum, Malaria, Lyme Disease, West Nile Virus, Chikungunya,
+LaCrosse Virus, Zika, Ehrlichia chaffeensis, and various forms of
+Typhus. Some cases, based on adherence to specific case definitions and
+subsequent lab results, are categorized as suspect, probable, confirmed,
+or unconfirmed. By examining case reports, public health professionals
+can better understand the prevalence, demographics, and regional
+variations of these diseases. This analysis aims to further inform the
+public health understanding of tick-borne disease in Georgia.
 
-## Questions/Hypotheses to be addressed
+## 3.1.2 Questions and Hypotheses to Be Addressed
 
-1.  How have vector-borne diseases (VBD) changed over time, and what
-    seasonal patterns or climatic factors influence these trends?
+This project seeks to explore temporal, spatial, and epidemiologic
+patterns of vector-borne disease (VBD) in Georgia using reported case
+data from 2019–2024. Specifically, it aims to address the following
+questions:
 
-2.  What is the distribution of vector-borne diseases across different
-    health districts, counties, and ZIP codes?
-
-3.  How complete is the data for vector-borne diseases, and how has data
+-   How complete is the data for vector-borne diseases, and how has data
     completeness evolved over time?
 
-4.  What is the timeline between key dates (onset, lab test, and date of
+-   What is the timeline between key dates (onset, lab test, and date of
     entry), excluding same-day onset and lab results, and how does this
@@
 
-5.  What trends exist in malaria travel cases over time, and which
-    countries are most frequently visited by affected individuals?
+-   Can we predict whether a confirmed tick-borne disease case will
+    result in a longer-than-typical diagnostic delay based on
+    demographic and geographic features (county of origin of the case
+    and Gini Index)?
 
-6.  What are the trends and changes observed in relative incidence of
-    different disease pathologies between 2019-2024?
-
-To cite other work (important everywhere, but likely happens first in
-introduction), make sure your references are in the bibtex file
-specified in the YAML header above and have the right bibtex key. Then
-you can include like this:
-
-Examples of reproducible research projects can for instance be found in
-[@mckay2020; @mckay2020a].
+    We hypothesized that tick-borne diseases would exhibit seasonal
+    trends corresponding with tick life cycles, that higher incidence
+    would be observed in areas with more natural and edge habitat, and
+    that case counts would correlate with both environmental conditions
+    and patterns of human interaction with tick habitats.
 
@@
 
-# Methods
+# 4. Methods
 
-**I. Uncleaned Data**
+## 4.1 Workflow Summarized
+
+-   Cleaned and prepared surveillance data on tick-borne disease cases
+    reported in Georgia from 2019–2024, including filtering for
+    confirmed positive cases and addressing issues with missing values
+    and inconsistent timelines.
+
+<!-- -->
+
+-   Conducted exploratory data analysis (EDA) to examine temporal
+    trends, geographic distribution, age patterns, diagnostic test
+    outcomes, and data completeness, using visualizations to inform
+    hypotheses and modeling decisions.
+
+-   Developed and evaluated multiple predictive models to assess whether
+    demographic and case-specific factors could explain delays in case
+    follow-up, using cross-validation and test/train splitting to assess
+    model quality and performance.
+
+## 4.2 Uncleaned Data
 
@@
 
-***II. Initial Data Cleaning Stages***
-
-Looked for Missing Data & General Trends
-
--   Identified Disease Burden Trends & Hotspot Counties
-
-<!-- -->
-
--   Generated a stacked bar graph to visualize case distribution across
-    regions. - Checked Age Distribution
-
-<!-- -->
-
--   Verified that no cases had missing age values in the cleaned
-    dataset.
-
-<!-- -->
-
--   Conducted an exploratory analysis of disease counts by age group.
-
--   Evaluated Data Completeness Over Time
-
--   Extracted onset date information to track missing data trends.
-
--   Created a plot distinguishing between blank and non-blank STATUS
-    entries over time.
-
--   Identified periods of high missingness, which may indicate data
-    entry errors or reporting gaps.
-
-Analyzed Timeline Between Key Case Events
-
--   Calculated time differences for:
-
-    -   Onset date (DOO) to first lab test date (LABDATE1)
-
-    -   First lab test (LABDATE1) to second lab test (LABDATE2)
-
-    -   Second lab test (LABDATE2) to case update date (UPDATEDATE)
-
-    -   Summed these values to determine the total case duration. -
-        Saved the processed dataset as a CSV file for further
-        analysis. - Created a formatted HTML table to present extracted
-        time intervals.
-
-Addressed Data Quality Issues
-
--   Identified negative time values in lab test dates, indicating data
-    entry errors
-
--   Implemented a filtering step to remove records with invalid date
-    sequences
-
-***III. Analysis Process***
-
--   Still determining this
-
-## Schematic of workflow
-
-Sometimes you might want to show a schematic diagram/figure that was not
-created with code (if you can do it with code, do it). @fig-schematic is
-an example of some - completely random/unrelated - schematic that was
-generated with Biorender. We store those figures in the `assets` folder.
-
-## Data aquisition
-
-The data used in this analysis was obtained through my internship with
-the Georgia Department of Health, in collaboration with the Centers for
-Disease Control and Prevention (CDC). Specifically, it is derived from
-the Nationally Notifiable Diseases Surveillance System (NNDSS) and
-focuses on tick-borne illnesses. This dataset provides key insights into
-the trends, distribution, and characteristics of reported cases,
-supporting public health efforts in disease monitoring and prevention.
-
-## Data import and cleaning
+## 4.3 Initial Data Cleaning
 
@@
 
-## Key Variables in Data Wrangling - Preparing for Analysis
+#### 4.3.1 Looked for Missing Data & General Trends
 
-To effectively present the cleaned dataset, key variables—including
-DISEASE, STATUS, COUNTY, GENDER, AGE, LABTEST1, and RESULT1—are
-displayed in a professionally formatted table using the kableExtra
-package. This enhances readability and facilitates trend analysis.
-Several visualizations are generated using ggplot2 to explore patterns
-in tick-borne diseases in Georgia between 2019 and 2024: Disease
-Distribution: A histogram illustrates the prevalence of different
-tick-borne diseases, highlighting the most commonly reported cases.
+-   Excluded cases with a missing or non-“CONFIRMED” STATUS to retain
+    only confirmed cases.
+-   Refined the dataset by retaining only confirmed “positive” test
+    results, excluding missing or ambiguous results, leaving 95 valid
+    cases.
+-   Additional filtering ensured correct identification of positive
+    cases while excluding non-positive values.
 
-**Rocky Mountain Spotted Fever (RMSF) Severity:**
+#### Addressed Data Quality Issues
 
-A box plot of TITER1 values (log-transformed for better visualization)
-provides insights into the variability of test results. RMSF severity
-patterns are considered analogous to Alpha-Gal Syndrome, a topic of
-interest, though not explicitly included in this dataset as it is not a
-reportable disease in Georgia.
+-   Identified negative time values between lab test dates, indicating
+    data entry errors.
+-   Implemented a filtering step to remove records with invalid date
+    sequences.
 
-```{r}
-library(here)
-library(knitr)
+## 4.4 Exploratory/Descriptive Analysis
 
-# Generate correct file path
-img_path <- here::here("results", "figures", "rmsf_titer_boxplot_log.png")
+#### 4.4.1 Identified Disease Burden Trends & Hotspot Counties
 
-# Display the image
-knitr::include_graphics(img_path)
+-   Generated a histogram to visualize the distribution of tick-borne
+    diseases in Georgia between 2019 and 2024.
+-   Created a bar chart highlighting the five most common diseases.
+-   Specifically analyzed Lyme Disease cases, focusing on test results.
 
-```
+#### 4.4.2 Checked Age Distribution
 
-**Most Common Tick-Borne Diseases:**
+-   Verified that no cases had missing age values in the cleaned
+    dataset.
+-   Conducted an exploratory analysis of disease counts by age group.
+
+#### 4.4.3 Evaluated Data Completeness Over Time
+
+-   Extracted onset date information to track missing data trends.
+-   Created a plot distinguishing between blank and non-blank STATUS
+    entries over time to identify periods of high missingness,
+    indicating possible data entry errors or reporting gaps.
+
+#### 4.4.4 Analyzed Timeline Between Key Case Events
+
+-   Calculated time differences for:
+    -   Onset date (DOO) to first lab test date (LABDATE1)
+    -   First lab test (LABDATE1) to second lab test (LABDATE2)
+    -   Second lab test (LABDATE2) to case update date (UPDATEDATE)
+-   Summed these values to determine the total case duration.
+-   Saved the processed dataset as a CSV file for further analysis.
+-   Created a formatted HTML table to present the extracted time
+    intervals.
+
+## 4.5 EDA Revelant Takeaways
+
+#### 4.5.1 Most Common Tick-Borne Diseases
 
@@
 
-```{r}
+```{r, echo=FALSE}
 library(here)
@@
 # Generate correct file path
-img_path <- here::here("results", "figures", "disease_histogram.png")
+img_path <- here::here("results", "supplemental-figures", "disease_histogram.png")
 
@@
 
-**Lyme Disease Analysis:**
+#### 4.5.2 Geographic Distribution
 
-Given its similarity to Alpha-Gal Syndrome
-(which is of interest to me and the topic of my capstone), Lyme Disease
-cases are analyzed separately. A count histogram compares positive,
-negative, and other test outcomes, providing insights into testing
-patterns and diagnosis rates.
+A grouped bar chart maps case counts across Georgia counties,
+categorized by disease type to analyze disease burden trends and
+identify hotspot counties. Due to the density of information, this
+visualization is refined to focus on the top five tick-borne diseases
+and their distribution across the 12 counties with the highest confirmed
+case counts.
 
-```{r}
-library(here)
-library(knitr)
-
-# Generate correct file path
-img_path <- here::here("results", "figures", "lyme_result_count_histogram2.png")
-
-# Display the image
-knitr::include_graphics(img_path)
-
-```
-
-**Geographic Distribution:**
-
-A grouped bar chart maps case counts across
-Georgia counties, categorized by disease type. Due to the density of
-information, this visualization is refined to focus on the top five
-tick-borne diseases and their distribution across the 12 counties with
-the highest confirmed case counts.
-```{r}
+```{r, echo=FALSE}
 library(here)
@@
 img_paths <- c(
-  here::here("results", "figures", "disease_county_plot.png"),
-  here::here("results", "figures", "top5_diseases_stacked_plot"),
-  here::here("results", "figures", "top5_disease_county_plot.png")
+  here::here("results", "supplemental-figures", "disease_county_plot.png"),
+  here::here("results", "supplemental-figures", "top5_diseases_stacked_plot.png"),
+  here::here("results", "supplemental-figures", "top5_disease_county_plot.png")
 )
@@
 
-To analyze disease burden trends and identify hotspot counties, a
-stacked bar graph is generated to visualize case distribution across
-different regions.
+#### 4.5.3 Data Completeness Over Time
 
-**Age Distribution Analysis**
+To evaluate how data completeness has evolved, the dataset is processed
+to extract onset date information. A visualization distinguishes between
+blank and non-blank entries in the STATUS column over time, helping to
+identify periods with high missingness. Spikes in blank values may
+indicate data entry errors, system issues, or reporting gaps that should
+be addressed before conducting further analysis.
 
-Age distribution is assessed as part of
-the initial data cleaning process, confirming that no cases in the
-cleaned dataset have missing age values. Additionally, an exploratory
-analysis examines disease counts by age group to identify potential
-patterns in disease prevalence.
-```{r}
+```{r, echo=FALSE}
 library(here)
@@
 # Generate correct file path
-img_path <- here::here("results", "figures", "disease_count_by_age.png")
+img_path <- here::here("results", "supplemental-figures", "status_completeness_plot.png")
 
@@
 
-**Data Completeness Over Time** 
-To evaluate how data completeness has
-evolved, the dataset is processed to extract onset date information. A
-visualization distinguishes between blank and non-blank entries in the
-STATUS column over time, helping to identify periods with high
-missingness. Spikes in blank values may indicate data entry errors,
-system issues, or reporting gaps that should be addressed before
-conducting further analysis.
+For more of my initial exploratory data analysis, please see the
+[Supplemental Figures]{.underline} section at the bottom of the report.
 
-```{r}
-library(here)
-library(knitr)
+#### 4.5.4 Timeline Analysis of Key Case Events
 
-# Generate correct file path
-img_path <- here::here("results", "figures", "status_completeness_plot.png")
+Given the frequent loss-to-follow-up observed in tick-borne illness
+reporting, an analysis is conducted to examine time gaps between
+critical dates, including:
 
-# Display the image
-knitr::include_graphics(img_path)
-```
+-   Onset date (DOO) to first lab test date (LABDATE1)
 
-**Timeline Analysis of Key Case Events** Given the frequent
-loss-to-follow-up observed in tick-borne illness reporting, an analysis
-is conducted to examine time gaps between critical dates, including:
+-   First lab test (LABDATE1) to second lab test (LABDATE2)
 
-**Onset date (DOO) to first lab test date (LABDATE1)**
-
-**First lab test (LABDATE1) to second lab test (LABDATE2)**
-
-**Second lab test (LABDATE2) to case update date (UPDATEDATE)**
+-   Second lab test (LABDATE2) to case update date (UPDATEDATE)
 
@@
 
-## Statistical analysis
+## 4.6 Statistical Analysis
 
-*Explain anything related to your statistical analyses.*
+To evaluate whether diagnostic delays could be predicted from case-level
+characteristics, a binary outcome variable (`delay_over_threshold`) was
+created to indicate whether the total time from symptom onset to final
+update exceeded 30 days. Predictor variables included patient age, sex,
+county of residence, and disease type. Age was normalized, and
+categorical variables were one-hot encoded.
+
+Three classification models—logistic regression, decision tree, and
+random forest—were trained using the `tidymodels`framework. Model
+performance was assessed using 5-fold cross-validation repeated three
+times. Metrics included accuracy and area under the receiver operating
+characteristic curve (AUC).
+
+Both basic bivariate analysis and statistical modeling was conducted.
+For the statistical modeling, three classification models were trained
+and evaluated using 5-fold cross-validation repeated 3 times.
+Performance was assessed using accuracy and area under the ROC curve
+(AUC).
+
+# 5. Results
+
+## 5.1 Basic statistical analysis
+
+For the first key question related to the disease trends over time (in
+years). In looking at the plot, it appears that there have been
+increases over time were observed in most of the diseases. However, the
+rate of increase was relatively slow as indicated by the lack of
+statistical significance of these trends (p\>0.5) which may indicate an
+attribution to the increase in reporting rather than an increase in
+overall disease burden. In the full analysis and second iteration of
+this, these trends will be compared year to year, comparing each year to
+the 2018 standard.
+
++-----------------+-----------+-----------+-----------+-----------+
+| term            | estimate  | s         | statistic | p.value   |
+|                 |           | td.error  |           |           |
+| \<chr\>         | \<dbl\>   |           | \<dbl\>   | \<dbl\>   |
+|                 |           | \<dbl\>   |           |           |
++:================+==========:+==========:+==========:+==========:+
+| ( Intercept)    | -2 3 2    | 1 57      | 1         | 0.1729390 |
+|                 | 8.5151515 | 3.052665  | .48025251 |           |
++-----------------+-----------+-----------+-----------+-----------+
+| YEAR            | 1.1515152 | 0.777581  | 1         | 0.1727716 |
+|                 |           |           | .48089418 |           |
++-----------------+-----------+-----------+-----------+-----------+
+| BABESIOSIS      | 0.4242424 | 5.484565  | 0         | 0.9400358 |
+|                 |           |           | .07735207 |           |
++-----------------+-----------+-----------+-----------+-----------+
+| DENGUE          | 4.8484848 | 6.364775  | 0         | 0.4656907 |
+|                 |           |           | .76176850 |           |
++-----------------+-----------+-----------+-----------+-----------+
+| EHRLICHIA       | 5.0000000 | 5.157889  | 0         | 0.3576794 |
+| CHAFFEENSIS     |           |           | .96938889 |           |
++-----------------+-----------+-----------+-----------+-----------+
+| EHRLICHIA       | -         | 6.364775  | 0         | 0.8604390 |
+| EWINGII         | 1.1515152 |           | .18092002 |           |
++-----------------+-----------+-----------+-----------+-----------+
+| LYME DISEASE    | 8.5606061 | 4.963745  | 1         | 0.1186814 |
+|                 |           |           | .72462659 |           |
++-----------------+-----------+-----------+-----------+-----------+
+| ROCKY MOUNTAIN  | 0.5757576 | 5.484565  | 0         | 0.9186958 |
+| SPOTTED FEVER   |           |           | .10497781 |           |
++-----------------+-----------+-----------+-----------+-----------+
+| WEST NILE (WNV) | 8.2272727 | 5.593721  | 1         | 0.1754208 |
+| INFECTION       |           |           | .47080497 |           |
++-----------------+-----------+-----------+-----------+-----------+
+
+9 rows
+
+```{r, echo=FALSE}
+library(here)
+library(knitr)
+
+# Generate correct file path
+img_path <- here::here("results", "supplemental-figures", "trend_over_time_by_disease.png")
+
+# Display the image
+knitr::include_graphics(img_path)
+```
+
+The most interesting finding from the initial modeling was in relation
+to the data completeness research question. In modeling the differences
+across time in whether a final STATUS was inputted rather than left
+blank, a general decrease was seen across the full 2018-2024 time scale.
+Specifically, completion rates of the reporting form were approximately
+50% in 2018 but dropped rapidly to 20% in 2019-- around which the rate
+of completeness has hovered since 2019. However, in linear modeling, the
+changes across time in the aggregate, p \> 0.05, meaning observed
+differences from year to year were not statistically significant.
+
+```{r, echo=FALSE}
+library(here)
+library(knitr)
+
+# Generate correct file path
+img_path <- here::here("results", "supplemental-figures", "completeness_over_time.png")
+
+# Display the image
+knitr::include_graphics(img_path)
+
+```
+
++----------+----------+----------+----------+----------+
+| term     | estimate | s        | s        | p.value  |
+|          |          | td.error | tatistic |          |
+| \<chr\>  | \<dbl\>  |          |          | \<dbl\>  |
+|          |          | \<dbl\>  | \<dbl\>  |          |
++:=========+=========:+=========:+=========:+=========:+
+| ( I      | 9 9      | 5 7      | 1.733920 | 0        |
+| n        | 3        | 2        |          | .1434648 |
+| tercept) | 5.267857 | 9.946482 |          |          |
++----------+----------+----------+----------+----------+
+| YEAR     | -        | 2.835202 | -        | 0        |
+|          | 4.910714 |          | 1.732051 | .1438108 |
++----------+----------+----------+----------+----------+
+
+2 rows
+
+On a positive note for the utilization of the reporting system and
+trends towards better tick-borne disease response, there has been
+statistically significant decreases in the time between onset and an
+initial lab visit/interaction. In 2018, the average time between these
+two occurrences ranged from approximately 150-520 days. With marked
+decreases, the 2024 data indicates that patients with tick-borne disease
+occurrences may be interacting with providers or lab testing in the same
+month, same week or even same day.
+
+```{r, echo=FALSE}
+library(here)
+library(knitr)
+
+# Generate correct file path
+img_path <- here::here("results", "supplemental-figures", "days_to_lab_over_time.png")
+
+# Display the image
+knitr::include_graphics(img_path)
+```
+
+## 5.2 Further Statistical Analysis
+
+To assess whether delays in disease follow-up could be predicted based
+on demographic or case-specific factors, a binary classification
+variable (`delay_over_threshold`) was created. This variable indicates
+whether the total case duration—from symptom onset to final
+update—exceeded 30 days. The outcome was treated as a binary factor (1 =
+delay \> 30 days, 0 = delay ≤ 30 days) and used as the target variable
+for modeling.
+
+Predictors included:
+
+-   AGE: Patient's age at time of case entry
+
+-   SEX: Reported gender
+
+-   COUNTY: County of residence
+
+-   DISEASE: Tick-borne disease type (e.g., Ehrlichiosis, Lyme disease,
+    RMSF)
+
+These predictors were selected based on their relevance to health access
+and geographic exposure patterns. All categorical variables were
+converted into dummy variables for modeling, and the age variable was
+normalized. Three classification models were trained and evaluated using
+the `tidymodel`framework:
+
+1.  Logistic Regression
+
+2.  Decision Tree
+
+3.  Random Forest
+
+To evaluate model performance, 5-fold cross-validation was repeated 3
+times on the training dataset. Model quality was assessed using both
+accuracy (overall correct classification rate) and area under the ROC
+curve (AUC), which measures the model's ability to discriminate between
+delayed and non-delayed cases.
+
+| Model               | Accuracy | ROC AUC |
+|---------------------|----------|---------|
+| Logistic Regression | 72.8%    | 0.65    |
+| Decision Tree       | 77.8%    | 0.61    |
+| Random Forest       | 78.1%    | 0.68    |
+
+The logistic regression model provided an interpretable baseline with
+modest discriminatory power (AUC = 0.65). The decision tree achieved
+higher accuracy but lower AUC, indicating possible overfitting or
+limited generalizability. The random forest model outperformed the other
+two in both metrics, achieving 78.1% accuracy and an AUC of 0.68, and
+was selected as the best-performing model. Given these results, the
+random forest model was selected for final evaluation on the test data
+set.
+
+```{r, echo=FALSE}
+library(here)
+library(knitr)
+
+knitr::include_graphics(here::here("results", "supplemental-figures", "roc_curve.png"))
+
+```
+
+## 5.3 Adding in Poverty Rate Data by County
+
+Now, I was curious. Reflecting back on the EDA analysis and the stark
+differences in county and cases, if the modeling of delay would better
+predicted if we merged county-level poverty rate data. This poverty rate
+data is reasonable, potential structural determinant of diagnostic delay
+due to persistent healthcare inequities. This variable was included in
+model training alongside demographic and disease characteristics.
+Preprocessing steps were updated accordingly, and models were
+re-evaluated using the same cross-validation and test evaluation
+procedures to assess any added predictive value.
+
+County-level poverty rate was evaluated as a predictor of whether a
+county had a high proportion of cases with diagnostic delays exceeding
+30 days. A random forest classification model using only poverty rate as
+a predictor achieved 88.4% accuracy, but its ability to discriminate
+between delayed and non-delayed counties was poor (ROC AUC = 0.46). This
+suggests that although the model often guessed the majority class
+correctly, it lacked the ability to meaningfully differentiate high-risk
+counties.
+
+Similarly, a linear regression model found no statistically significant
+association between poverty rate and delay rate (p = 0.87, β = 0.0049).
+The random forest regression model explained less than 4% of the
+variance in delay rates (R² = 0.038) with a relatively high prediction
+error (RMSE = 0.37).
+
++----------------+----------------+----------------+----------------+
+|                |                |                |                |
++----------------+----------------+----------------+----------------+
+|                |                |                |                |
++----------------+----------------+----------------+----------------+
 
@@
 
-# Results
+# 6. Discussion
 
-## Exploratory/Descriptive analysis
+## 6.1 Summary and Interpretation
 
-*Use a combination of text/tables/figures to explore and describe your
-data. Show the most important descriptive results here. Additional ones
-should go in the supplement. Even more can be in the R and Quarto files
-that are part of your project.*
+The random forest model for the modeling of delay as an outcome all
+predictors selected from cross-validation was refitted to the full
+training data and evaluated on the reserved 20% test dataset to assess
+generalizability. On this unseen data, the model achieved an accuracy of
+74.7%, meaning nearly three-quarters of tick-borne disease cases were
+correctly classified as having experienced delayed (\>30 day) or
+non-delayed follow-up. The model’s ability to discriminate between
+delayed and non-delayed cases was moderate, with an area under the
+receiver operating characteristic (ROC) curve of 0.688, closely
+mirroring performance observed during cross-validation.
 
-Note the loading of the data providing a **relative** path using the
-`../../` notation. (Two dots means a folder up). You never want to
-specify an **absolute** path like `C:\ahandel\myproject\results\`
-because if you share this with someone, it won't work for them since
-they don't have that path. You can also use the `here` R package to
-create paths. See examples of that below. I generally recommend the
-`here` package.
+The ROC curve showed consistent separation between sensitivity and
+1-specificity, further supporting model validity. These results suggest
+that case characteristics such as age, sex, county, and disease type
+contain meaningful predictive information regarding follow-up delays in
+surveillance data.Such insights can inform improvements in case
+monitoring, triaging, and outreach in counties or patient groups with
+higher risk for delays.
 
-## Basic statistical analysis
+County-level poverty rate was evaluated as a predictor of the proportion
+of cases with diagnostic delays exceeding 30 days. A linear regression
+model indicated no statistically significant relationship between
+poverty rate and delay rate (p = 0.87). The estimated effect size was
+minimal (β = 0.0049), and the 95% confidence interval included zero.
 
-## Full analysis
+A random forest regression model using the same predictor resulted in
+low predictive performance. The R² was 0.038, indicating that poverty
+rate explained less than 4% of the variance in delay rate across
+counties. The RMSE was 0.37, reflecting high residual error in
+predicting delay proportions.
 
-*Use one or several suitable statistical/machine learning methods to
-analyze your data and to produce meaningful figures, tables, etc. This
-might again be code that is best placed in one or several separate R
-scripts that need to be well documented. You want the code to produce
-figures and data ready for display as tables, and save those. Then you
-load them here.*
+In comparison, the earlier classification models using individual-level
+predictors (e.g., age, sex, disease type, county) performed considerably
+better in both accuracy and AUC, suggesting that county-level poverty
+alone does not substantially contribute to explaining or predicting
+diagnostic delays in this dataset.
+
+### What this means:
+
+-   The poverty-only model had higher accuracy, but a lower ROC AUC.
+
+-   Accuracy tells us how often the model got it right — so poverty
+    alone seems to classify “delay” correctly most of the time.
+
+-   However, the low ROC AUC suggests that the model struggles to rank
+    cases by risk — in other words, it doesn’t distinguish well between
+    high and low risk of delay.
+
+-   The all-predictors model performed more balanced, with a decent
+    accuracy and a stronger ROC AUC, meaning it did a better
+    job discriminating between delayed and non-delayed cases.
+
++----------------+----------------+----------------+----------------+
+| Model Type     | Predictors     | Accuracy       | ROC AUC        |
+|                | Used           |                |                |
++================+================+================+================+
+| Random Forest  | Age, Sex,      | 78.1%          | 0.68           |
+|                | County,        |                |                |
+|                | Disease        |                |                |
++----------------+----------------+----------------+----------------+
+| Random         | Poverty Rate   | 88.4%          | 0.46           |
+| Forest -       | By County      |                |                |
+| Poverty Rate   |                |                |                |
+| Only           |                |                |                |
++----------------+----------------+----------------+----------------+
+
+## 6.2 Strengths and Limitations
+
+This analysis leveraged a comprehensive dataset spanning six years and
+multiple counties across Georgia, allowing for exploration of both
+individual- and community-level patterns in diagnostic delay for
+tick-borne disease cases. The geographic breadth of the data is a key
+strength, enabling comparisons across rural and urban areas and
+supporting more generalizable insights. Additionally, the use of several
+modeling strategies—including logistic regression, decision trees, and
+random forests—provided a well-rounded evaluation of predictor
+importance and model performance. The inclusion of both cross-validation
+and a holdout test set added rigor and reduced the risk of overfitting.
+
+Another limitation is the use of county-level poverty rate as a proxy
+for individual socioeconomic status (SES). While poverty data at the
+county level offers a broad socioeconomic context, it can mask
+significant within-county income disparities. In counties with wide
+economic variation, aggregate poverty rates may not accurately represent
+the lived experience or access to healthcare of individuals affected by
+tick-borne disease.
+
+Furthermore, individual exposure risk is often tightly linked to SES.
+For instance, people in outdoor labor roles—such as landscaping,
+farming, or construction—may face greater tick exposure than those
+working indoors, regardless of the county’s average income level. These
+more granular, individual-level occupational or behavioral factors were
+not captured in this dataset, which limits the precision of predictive
+modeling based solely on geography and aggregated socioeconomic
+indicators.
+
+However, several limitations should be acknowledged. Inconsistencies in
+reporting across cases and counties required extensive data cleaning,
+which ultimately reduced the dataset to a much smaller number of usable
+observations. This constrained sample size may have limited model
+performance and generalizability. The dataset also lacked clinical and
+healthcare access variables—such as symptom severity, insurance status,
+or provider responsiveness—which likely influence diagnostic delays.
+Without these contextual details, the models may have missed key drivers
+of delay. Finally, the poverty rate was measured at the county level,
+which may not reflect intra-county disparities or the lived experience
+of patients.
+
+# 7. Conclusions
+
+-   Individual-level characteristics such as age, sex, disease type, and
+    county were moderately effective in predicting whether a tick-borne
+    disease case experienced a diagnostic delay of over 30 days. The
+    best-performing model (random forest) achieved 78.1% accuracy and an
+    AUC of 0.68, indicating meaningful—though not perfect—predictive
+    value.
+
+-   County-level poverty rate alone was not a strong predictor of
+    diagnostic delay. Both linear regression and random forest
+    regression showed weak associations and low explanatory power,
+    suggesting that broader socioeconomic indicators may not capture the
+    nuanced drivers of individual-level follow-up delays.
+
+-   Data quality and completeness were major challenges: extensive
+    cleaning was required, and missing or inconsistent entries
+    significantly reduced the number of usable cases. This limitation
+    underscores the need for more standardized reporting practices in
+    public health surveillance systems.
 
@@
 
-# Discussion
+# Supplemental Figures
 
-## Summary and Interpretation
+A full set of exploratory and supplemental visualizations—used to guide
+data cleaning, quality checks, and hypothesis generation—are available
+in a separate folder.
 
-*Summarize what you did, what you found and what it means.*
+Please refer to the following file for all supplemental figures:
 
-## Strengths and Limitations
+File: `supplemental-figures.zip`
 
-*Discuss what you perceive as strengths and limitations of your
-analysis.*
-
-## Conclusions
-
-*What are the main take-home messages?*
-
-*Include citations in your Rmd file using bibtex, the list of references
-will automatically be placed at the end*
-
-{{< pagebreak >}}
+This archive includes: - Disease distribution histograms - County-level
+maps - Test result breakdowns - Titer plots - Age distribution and data
+completeness trends that I did in the exploratory data analysis but
+opted to not include in my analysis per my reviewers' recommendations
+that I narrow my scope!
 

### 2025-04-21 code/processing-code/processing-code-tickcases.qmd
Update
Reviewer - I need help with why on Docx my numbering gets all weird -- the numbering looks fine on my end on R

@@
 # Create histogram of disease counts
-disease_histogram <- ggplot(cleaned_data3, aes(x = DISEASE)) +
+disease_plot <- ggplot(cleaned_data3, aes(x = DISEASE)) +
   geom_bar(fill = "#2C3E50", color = "white") +  # Dark blue bars with white edges
@@
 # Save the plot
-ggsave(filename = save_plot_location, plot = disease_histogram, width = 12, height = 6, dpi = 300)
+ggsave(filename = save_plot_location, plot = disease_plot, width = 12, height = 6, dpi = 300)
 ```

### 2025-04-21 products/manuscript/Manuscript.qmd
Update
Reviewer - I need help with why on Docx my numbering gets all weird -- the numbering looks fine on my end on R

@@
 
-+-----------------+-----------+-----------+-----------+-----------+
++-----------+-----------+-----------+-----------+-----------+
 | term      | estimate  | s         | statistic | p.value   |
@@
 |           |           | \<dbl\>   |           |           |
-+:================+==========:+==========:+==========:+==========:+
-| ( Intercept)    | -2 3 2    | 1 57      | 1         | 0.1729390 |
-|                 | 8.5151515 | 3.052665  | .48025251 |           |
-+-----------------+-----------+-----------+-----------+-----------+
++:==========+==========:+==========:+==========:+==========:+
+| (         | -2 3 2    | 1 57      | 1         | 0.1729390 |
+| I         | 8.5151515 | 3.052665  | .48025251 |           |
+| ntercept) |           |           |           |           |
++-----------+-----------+-----------+-----------+-----------+
 | YEAR      | 1.1515152 | 0.777581  | 1         | 0.1727716 |
 |           |           |           | .48089418 |           |
-+-----------------+-----------+-----------+-----------+-----------+
-| BABESIOSIS      | 0.4242424 | 5.484565  | 0         | 0.9400358 |
-|                 |           |           | .07735207 |           |
-+-----------------+-----------+-----------+-----------+-----------+
++-----------+-----------+-----------+-----------+-----------+
+| B         | 0.4242424 | 5.484565  | 0         | 0.9400358 |
+| ABESIOSIS |           |           | .07735207 |           |
++-----------+-----------+-----------+-----------+-----------+
 | DENGUE    | 4.8484848 | 6.364775  | 0         | 0.4656907 |
 |           |           |           | .76176850 |           |
-+-----------------+-----------+-----------+-----------+-----------+
++-----------+-----------+-----------+-----------+-----------+
 | EHRLICHIA | 5.0000000 | 5.157889  | 0         | 0.3576794 |
-| CHAFFEENSIS     |           |           | .96938889 |           |
-+-----------------+-----------+-----------+-----------+-----------+
+| C         |           |           | .96938889 |           |
+| H         |           |           |           |           |
+| AFFEENSIS |           |           |           |           |
++-----------+-----------+-----------+-----------+-----------+
 | EHRLICHIA | -         | 6.364775  | 0         | 0.8604390 |
 | EWINGII   | 1.1515152 |           | .18092002 |           |
-+-----------------+-----------+-----------+-----------+-----------+
-| LYME DISEASE    | 8.5606061 | 4.963745  | 1         | 0.1186814 |
-|                 |           |           | .72462659 |           |
-+-----------------+-----------+-----------+-----------+-----------+
-| ROCKY MOUNTAIN  | 0.5757576 | 5.484565  | 0         | 0.9186958 |
-| SPOTTED FEVER   |           |           | .10497781 |           |
-+-----------------+-----------+-----------+-----------+-----------+
-| WEST NILE (WNV) | 8.2272727 | 5.593721  | 1         | 0.1754208 |
-| INFECTION       |           |           | .47080497 |           |
-+-----------------+-----------+-----------+-----------+-----------+
++-----------+-----------+-----------+-----------+-----------+
+| LYME      | 8.5606061 | 4.963745  | 1         | 0.1186814 |
+| DISEASE   |           |           | .72462659 |           |
++-----------+-----------+-----------+-----------+-----------+
+| ROCKY     | 0.5757576 | 5.484565  | 0         | 0.9186958 |
+| MOUNTAIN  |           |           | .10497781 |           |
+| SPOTTED   |           |           |           |           |
+| FEVER     |           |           |           |           |
++-----------+-----------+-----------+-----------+-----------+
+| WEST NILE | 8.2272727 | 5.593721  | 1         | 0.1754208 |
+| (WNV)     |           |           | .47080497 |           |
+| INFECTION |           |           |           |           |
++-----------+-----------+-----------+-----------+-----------+
 
@@
 
-+----------+----------+----------+----------+----------+
-| term     | estimate | s        | s        | p.value  |
-|          |          | td.error | tatistic |          |
-| \<chr\>  | \<dbl\>  |          |          | \<dbl\>  |
-|          |          | \<dbl\>  | \<dbl\>  |          |
-+:=========+=========:+=========:+=========:+=========:+
-| ( I      | 9 9      | 5 7      | 1.733920 | 0        |
-| n        | 3        | 2        |          | .1434648 |
-| tercept) | 5.267857 | 9.946482 |          |          |
-+----------+----------+----------+----------+----------+
-| YEAR     | -        | 2.835202 | -        | 0        |
-|          | 4.910714 |          | 1.732051 | .1438108 |
-+----------+----------+----------+----------+----------+
++--------+--------+--------+--------+--------+
+| term   | e      | s t    | s t    | p      |
+|        | s      | d      | a      | .value |
+| \      | timate | .error | tistic |        |
+| <chr\> |        |        |        | \      |
+|        | \      | \      | \      | <dbl\> |
+|        | <dbl\> | <dbl\> | <dbl\> |        |
++:=======+=======:+=======:+=======:+=======:+
+| ( I n  | 9 9 3  | 5 7 2  | 1      | 0 .    |
+| t      | 5      | 9      | .      | 1      |
+| e      | .      | .      | 733920 | 434648 |
+| rcept) | 267857 | 946482 |        |        |
++--------+--------+--------+--------+--------+
+| YEAR   | -   4  | 2      | -   1  | 0 .    |
+|        |     .  | .      |     .  | 1      |
+|        | 910714 | 835202 | 732051 | 438108 |
++--------+--------+--------+--------+--------+
 
@@
 
-+----------------+----------------+----------------+----------------+
++--------------+--------------+--------------+--------------+
 |              |              |              |              |
-+----------------+----------------+----------------+----------------+
++--------------+--------------+--------------+--------------+
 |              |              |              |              |
-+----------------+----------------+----------------+----------------+
++--------------+--------------+--------------+--------------+
 
@@
 
-+----------------+----------------+----------------+----------------+
++--------------+--------------+--------------+--------------+
 | Model Type   | Predictors   | Accuracy     | ROC AUC      |
 |              | Used         |              |              |
-+================+================+================+================+
-| Random Forest  | Age, Sex,      | 78.1%          | 0.68           |
-|                | County,        |                |                |
++==============+==============+==============+==============+
+| Random       | Age, Sex,    | 78.1%        | 0.68         |
+| Forest       | County,      |              |              |
 |              | Disease      |              |              |
-+----------------+----------------+----------------+----------------+
++--------------+--------------+--------------+--------------+
 | Random       | Poverty Rate | 88.4%        | 0.46         |
@@
 | Only         |              |              |              |
-+----------------+----------------+----------------+----------------+
++--------------+--------------+--------------+--------------+
 

### 2025-04-26 project-review-template-nasir.qmd
Review 5_Muhammad Nasir

---
title: Project Review 
author: Hope Grismer
date: date-modified
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---

# Overview

Title of project:Analysis of Tick-Bourne Disease Cases Reported to DPH Between 2019-2024

Name of project author(s):Hope Grismer

Name of project reviewer: Muhammad Nasir

# Instructions

Write your comments and feedback below for each section/component of the project. The goal should be to help the author improve their project. Make comments as constructive and actionable as possible. You can provide both criticism and praise.

For each component, pick one summary statement by deleting the ones that do not apply and keeping only the one that you think most closely summarizes a given component.

Make sure your final document compiles/renders into a readable, well-formatted html document.

Delete any sections/text of this template that are not part of your final review document. (Including these instructions.)

## Background, Context and Motivation

### Feedback and Comments

The introduction part provides a clear description including background information about the diseases, which is very important for readers who have limited knowledge about Tick-borne illnesses (TBDs), tranmission of the diseases, and the revelance of the topic. 
The only thing missing from this part is citation. I think it is important to provide some sources in the background information.

### Summary assessment (PICK ONE, DELETE THE OTHERS)

* strong contextualization and motivation

## Question description

### Feedback and Comments

Research Questions and hypotheses are discribed clearly relevan with knowldge gaps provided in the context section.

### Summary assessment
* question/hypotheses fully clear

## Data description

### Feedback and Comments

The data source is provided. However, I did not see codebook.

### Summary assessment
* source and overall structure of data well explained

## Data wrangling and exploratory analysis

### Feedback and Comments
The data was explored very good with notes in every steps of the process. This is super helpfull. However, there is no explanation about the plots/ figures. It would be better to provide.
I like the reasons of performing particular data analysis explained.

### Summary assessment

* some weaknesses in wrangling and exploratory component

## Appropriateness of Analysis
Were the analysis methods appropriate for the data? Was the analysis done properly? Were different components of the analysis (e.g. performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?

### Feedback and Comments

- It is clear that the study performed three different methods: logistic regression, desicion tree and random forest. Random forest performed best in this dataset. 
- I am a little bit worry about perfoming linear model for day_to_lab ~ year. We need to make sure that all assumptions of linear regression met. The 2019 point looks faraway, it might became an ourliner and seems influence the model so much. If we pay more attention for the next years, it shows increasing trend until 2023 before dicrease to 2024.

### Summary assessment

* strong and reasonable analysis

## Presentation

### Feedback and Comments
- The overall presentation very good, using different plots and table to visualize the result. The plot for ROC is very useful.

- I am a bit comfused about the table in 5.1. Which analyisi the table came from? Is it linar regression? why does the table shows type of diseases?

### Summary assessment
* results are presented ok, with room for improvement

## Discussion/Conclusions

### Feedback and Comments

The finding is well discussed. However, it will be much better if the result compared with existing/ related studies to support. Strenght and limitations are addressed well.

### Summary assessment

* minor parts wrong, missing or unclear

## Further comments

_Add any other comments regarding the different aspects of the project here. Write anything you think can help your classmate improve their project._

# Overall project content evaluation

## Structure
Is the project well structured? Are files in well labeled folders? Do files have reasonable names? Are all "junk" files not needed for analysis/reproduction removed? By just looking at files and folders, can you get an idea of how things fit together?

### Feedback and Comments

The project structure is good. However, readme.md files need to be updated. Some folders/ files need to be removed.

### Summary assessment

* mostly clear, but some confusing parts (e.g. useless files, things in the wrong folders)

## Documentation 
How well is the project documented? Are you able to understand each step of the whole analysis, each decision that was made, and each line of code? Is enough information provided as comments in code or as part of Rmd files?

### Feedback and Comments

output such as figures and tables are well documented in a folder.

### Summary assessment

* fully and well documented

## Reproducibility
Are all results fully reproducible? Is documentation provided which clearly explains how to reproduce things, and does it work without the need for any manual intervention? Are you able to re-run the whole analysis without having to do manual interventions/edits?

### Feedback and Comments

- I can run most part of the code. I only got error in library (dplyr), I needed to install with defendency. 
- The manuscript cannot be render due to reference missing.

### Summary assessment

* small parts not reproducible or required manual intervention

## Thoroughness
How thorough was the overall study? Were alternatives (e.g. different ways of processing the data or different models) considered? Were alternatives discussed? Were the questions/hypotheses fully and thoroughly addressed?

### Feedback and Comments

WRITE YOUR FEEDBACK HERE

### Summary assessment

* strong level of thorougness

## Further comments

1.	I got issue when library (dplyr) in the packages section of “processing-code-tickcases.qdm”. I need to run this code  install.packages("dplyr", dependencies = FALSE, type = "binary"). I don’t know why it happened, could be from my side. 
2.	It is better to not install/ library package in the chunk of your code, but do it in specific place, check : 
-	 processing-code-tickcases.qdm  line 251-254 , line 358-361, line 406- 410, line 488-489, line 528-530, line 568-573, line 620-626. 
3.	Issues in “statistical-analysis-gini-index.qdm”
-	Packages are install/ library in code chunk: line 10-13, 
-	Line 20, the path/ or data location is not exist. 
-	If the file is important or part of the project, we need to provide the dataset. However, if this analysis is not an important part of the project, then we can remove it. 
4.	Issue in “statistical-analysis-poverty-rate.qdm”
-	Packages are install/ library in code chunk: line 2-4 
-	This is the same with previous file, if this is not an important part of the analysis in the project, we can remove it. 
5.	Issues in “statistical-analysis.qdm”
-	Packages are install/ library in code chunk: line 118-122, line 232, line 258
-	It would be better to put description/ interpretation in each figure. 
-	In the study description, it says that the study will use data from 2019 to 2024, but in the data completeness, the figure shows the data from 2018 (see part 3. completeness)
-	For the fiu
6.	Issue in manuscript.qdm 
-	4.5.1 Most common tick-borne disease: In this section, it says five most frequent;y reposrted cases, in my understanding, the chart will only show five TBDs. 
-	Some label value in TBDs by county chart are not in appropriate places. 
-	There are three same figures in place for TBDs cases by county. I think these figures are same. 
-	Empty tables  line 646-650 
7.	Readme.md files need to be updated.

### 2025-04-26 products/manuscript/Manuscript.qmd

@@
 
-+-----------+-----------+-----------+-----------+-----------+
-| term      | estimate  | s         | statistic | p.value   |
-|           |           | td.error  |           |           |
-| \<chr\>   | \<dbl\>   |           | \<dbl\>   | \<dbl\>   |
-|           |           | \<dbl\>   |           |           |
-+:==========+==========:+==========:+==========:+==========:+
-| (         | -2 3 2    | 1 57      | 1         | 0.1729390 |
-| I         | 8.5151515 | 3.052665  | .48025251 |           |
-| ntercept) |           |           |           |           |
-+-----------+-----------+-----------+-----------+-----------+
-| YEAR      | 1.1515152 | 0.777581  | 1         | 0.1727716 |
-|           |           |           | .48089418 |           |
-+-----------+-----------+-----------+-----------+-----------+
-| B         | 0.4242424 | 5.484565  | 0         | 0.9400358 |
-| ABESIOSIS |           |           | .07735207 |           |
-+-----------+-----------+-----------+-----------+-----------+
-| DENGUE    | 4.8484848 | 6.364775  | 0         | 0.4656907 |
-|           |           |           | .76176850 |           |
-+-----------+-----------+-----------+-----------+-----------+
-| EHRLICHIA | 5.0000000 | 5.157889  | 0         | 0.3576794 |
-| C         |           |           | .96938889 |           |
-| H         |           |           |           |           |
-| AFFEENSIS |           |           |           |           |
-+-----------+-----------+-----------+-----------+-----------+
-| EHRLICHIA | -         | 6.364775  | 0         | 0.8604390 |
-| EWINGII   | 1.1515152 |           | .18092002 |           |
-+-----------+-----------+-----------+-----------+-----------+
-| LYME      | 8.5606061 | 4.963745  | 1         | 0.1186814 |
-| DISEASE   |           |           | .72462659 |           |
-+-----------+-----------+-----------+-----------+-----------+
-| ROCKY     | 0.5757576 | 5.484565  | 0         | 0.9186958 |
-| MOUNTAIN  |           |           | .10497781 |           |
-| SPOTTED   |           |           |           |           |
++----------+----------+----------+----------+----------+
+| term     | estimate | s        | s        | p.value  |
+|          |          | td.error | tatistic |          |
+| \<chr\>  | \<dbl\>  |          |          | \<dbl\>  |
+|          |          | \<dbl\>  | \<dbl\>  |          |
++:=========+=========:+=========:+=========:+=========:+
+| ( I      | -2 3 2   | 1 57     | 1        | 0        |
+| n        | 8        | 3.052665 | .        | .1729390 |
+| tercept) | .5151515 |          | 48025251 |          |
++----------+----------+----------+----------+----------+
+| YEAR     | 1        | 0.777581 | 1        | 0        |
+|          | .1515152 |          | .        | .1727716 |
+|          |          |          | 48089418 |          |
++----------+----------+----------+----------+----------+
+| B        | 0        | 5.484565 | 0        | 0        |
+| A        | .4242424 |          | .        | .9400358 |
+| BESIOSIS |          |          | 07735207 |          |
++----------+----------+----------+----------+----------+
+| DENGUE   | 4        | 6.364775 | 0        | 0        |
+|          | .8484848 |          | .        | .4656907 |
+|          |          |          | 76176850 |          |
++----------+----------+----------+----------+----------+
+| E        | 5        | 5.157889 | 0        | 0        |
+| HRLICHIA | .0000000 |          | .        | .3576794 |
+| C H      |          |          | 96938889 |          |
+| A        |          |          |          |          |
+| FFEENSIS |          |          |          |          |
++----------+----------+----------+----------+----------+
+| E        | -   1    | 6.364775 | 0        | 0        |
+| HRLICHIA | .1515152 |          | .        | .8604390 |
+| EWINGII  |          |          | 18092002 |          |
++----------+----------+----------+----------+----------+
+| LYME     | 8        | 4.963745 | 1        | 0        |
+| DISEASE  | .5606061 |          | .        | .1186814 |
+|          |          |          | 72462659 |          |
++----------+----------+----------+----------+----------+
+| ROCKY    | 0        | 5.484565 | 0        | 0        |
+| MOUNTAIN | .5757576 |          | .        | .9186958 |
+| SPOTTED  |          |          | 10497781 |          |
 | FEVER    |          |          |          |          |
-+-----------+-----------+-----------+-----------+-----------+
-| WEST NILE | 8.2272727 | 5.593721  | 1         | 0.1754208 |
-| (WNV)     |           |           | .47080497 |           |
-| INFECTION |           |           |           |           |
-+-----------+-----------+-----------+-----------+-----------+
++----------+----------+----------+----------+----------+
+| WEST     | 8        | 5.593721 | 1        | 0        |
+| NILE     | .2272727 |          | .        | .1754208 |
+| (WNV)    |          |          | 47080497 |          |
+| I        |          |          |          |          |
+| NFECTION |          |          |          |          |
++----------+----------+----------+----------+----------+
 
@@
 
-+--------+--------+--------+--------+--------+
-| term   | e      | s t    | s t    | p      |
-|        | s      | d      | a      | .value |
-| \      | timate | .error | tistic |        |
-| <chr\> |        |        |        | \      |
-|        | \      | \      | \      | <dbl\> |
-|        | <dbl\> | <dbl\> | <dbl\> |        |
-+:=======+=======:+=======:+=======:+=======:+
-| ( I n  | 9 9 3  | 5 7 2  | 1      | 0 .    |
-| t      | 5      | 9      | .      | 1      |
-| e      | .      | .      | 733920 | 434648 |
-| rcept) | 267857 | 946482 |        |        |
-+--------+--------+--------+--------+--------+
-| YEAR   | -   4  | 2      | -   1  | 0 .    |
-|        |     .  | .      |     .  | 1      |
-|        | 910714 | 835202 | 732051 | 438108 |
-+--------+--------+--------+--------+--------+
++-------+-------+-------+-------+-------+
+| term  | e s   | s t d | s t a | p     |
+|       | t     | .     | t     | .     |
+| \     | imate | error | istic | value |
+| \<    |       |       |       |       |
+| chr\> | \     | \     | \     | \     |
+|       | \<    | \<    | \<    | \<    |
+|       | dbl\> | dbl\> | dbl\> | dbl\> |
++:======+======:+======:+======:+======:+
+| ( I n | 9 9 3 | 5 7 2 | 1 .   | 0 . 1 |
+| t e   | 5 .   | 9 .   | 7     | 4     |
+| r     | 2     | 9     | 33920 | 34648 |
+| cept) | 67857 | 46482 |       |       |
++-------+-------+-------+-------+-------+
+| YEAR  | -   4 | 2 .   | -   1 | 0 . 1 |
+|       |     . | 8     |     . | 4     |
+|       |     9 | 35202 |     7 | 38108 |
+|       | 10714 |       | 32051 |       |
++-------+-------+-------+-------+-------+
 
@@
 
-+--------------+--------------+--------------+--------------+
++-------------+-------------+-------------+-------------+
 |             |             |             |             |
-+--------------+--------------+--------------+--------------+
++-------------+-------------+-------------+-------------+
 |             |             |             |             |
-+--------------+--------------+--------------+--------------+
++-------------+-------------+-------------+-------------+
 
@@
 
-+--------------+--------------+--------------+--------------+
++-------------+-------------+-------------+-------------+
 | Model Type  | Predictors  | Accuracy    | ROC AUC     |
 |             | Used        |             |             |
-+==============+==============+==============+==============+
++=============+=============+=============+=============+
 | Random      | Age, Sex,   | 78.1%       | 0.68        |
@@
 |             | Disease     |             |             |
-+--------------+--------------+--------------+--------------+
-| Random       | Poverty Rate | 88.4%        | 0.46         |
-| Forest -     | By County    |              |              |
-| Poverty Rate |              |              |              |
-| Only         |              |              |              |
-+--------------+--------------+--------------+--------------+
++-------------+-------------+-------------+-------------+
+| Random      | Poverty     | 88.4%       | 0.46        |
+| Forest -    | Rate By     |             |             |
+| Poverty     | County      |             |             |
+| Rate Only   |             |             |             |
++-------------+-------------+-------------+-------------+
 

### 2025-04-26 Norris-Review.qmd

---
title: Hope Grismer Project Reivew
author: Connor Norris
date: 2025-04-25
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
---

# Overview

Title of project: Analysis of Tick-Bourne Disease Cases Reported to DPH Between 2019-2024

Name of project author(s): Hope Grismer

Name of project reviewer: Connor Norris

# Specific project content evaluation

## Background, Context and Motivation
How well is the context of the project described? Is a comprehensive background, including summary of previous/related work given? Is the project well placed into the context of existing work (including proper referencing of existing work). Is it clear why the project was undertaken and what new information it hopes to provide?

### Feedback and Comments

Your background is very well written and is definitely comprehensive. However, almost everything you have in the background needs a source cited.

### Summary assessment (PICK ONE, DELETE THE OTHERS)

* strong contextualization and motivation

## Question description
How well and clear are the question(s)/hypotheses the project aims to address described? Is it clear how the questions relate to the data?

### Feedback and Comments

I think you explain your rationale for doing this analysis very well and put it in the context of the information you provide in your background. However, I don't know what specific questions that you have because of that rationale and what you expect the answers of those questions to be. Take a couple sentences in the last paragraph of your intro to explain what you did with this analysis, big picture.

### Summary assessment
* question/hypotheses fully clear

## Data description
How well is the data overall described? Is the source provided? Is a codebook or other meta-information available that makes it clear what the data is?

### Feedback and Comments

First, move the data description to the beginning of the methods. Putting it in between the background and your research questions makes your questions not as immediately clear to readers. That being said, I think this section is well done. The only thing I would change is to clarify what 'cases' mean at the very beginning of your dataset description. Does the data only include cases of tickborne disease, or is it any reportable disease?

### Summary assessment

* source and overall structure of data well explained

## Data wrangling and exploratory analysis
How well is the data cleaned/processed and explored? Are all steps reasonable and well explained? Are alternatives discussed and considered? Are meaningful exploratory results shown (e.g. in the supplementary materials)?

### Feedback and Comments

I think you almost have too much information in your data wrangling section of the manuscript, but I don't think that's a bad thing! I think your exploratory data analysis is also super thorough.

### Summary assessment

* essentially no weaknesses in wrangling and exploratory component

## Appropriateness of Analysis
Were the analysis methods appropriate for the data? Was the analysis done properly? Were different components of the analysis (e.g. performance measure, variable selection, data pre-processing, model evaluation) done in the best way possible and explained well?

### Feedback and Comments

I think that all of your choices for how you did your analysis were well made, and I appreciated how you used multiple models.

### Summary assessment

* strong and reasonable analysis

## Presentation
How well are results presented? Are tables and figures easy to read and understand? Are the main figures/tables publication level quality?

### Feedback and Comments

For your tables, I like the information you put in all of them, but I don't think the format you put them in translates well to the Word document (the spacing of words is weird), and I don't think you would be able to reproduce them easily if you had to tweak your analysis. You have good tables in your code, though, just save them and use them in the manuscript. I think for the most part, your figures are well done. The only one I would reconsider is the vector-borne diseases over time by disease figure. It's hard to see individual trends. I think your figure where you plot blanks versus text in the STATUS column by month is super cool.

### Summary assessment

* results are presented ok, with room for improvement

## Discussion/Conclusions
Are the study findings properly discussed? Are strengths and limitations acknowledged? Are findings interpreted properly?

### Feedback and Comments

I think your discussion is well done, and I appreciate how you talked about your limitations.

### Summary assessment

* strong, complete and clear discussion

## Further comments

# Overall project content evaluation
Evaluate overall features of the project  by filling in the sections below.

## Structure
Is the project well structured? Are files in well labeled folders? Do files have reasonable names? Are all "junk" files not needed for analysis/reproduction removed? By just looking at files and folders, can you get an idea of how things fit together?

### Feedback and Comments

Your project directory is overall well structured, just make sure you edit or delete the readme files across the directory.

### Summary assessment

* mostly clear, but some confusing parts (e.g. useless files, things in the wrong folders)

## Documentation 
How well is the project documented? Are you able to understand each step of the whole analysis, each decision that was made, and each line of code? Is enough information provided as comments in code or as part of Rmd files?

### Feedback and Comments

I think you document your code well - it was pretty easy to read through it all. In the readme file of the analysis-code folder, I would briefly explain what each file does.

### Summary assessment

* fully and well documented

## Reproducibility
Are all results fully reproducible? Is documentation provided which clearly explains how to reproduce things, and does it work without the need for any manual intervention? Are you able to re-run the whole analysis without having to do manual interventions/edits?

### Feedback and Comments

I was able to run your code with no problem. However, I was unable to load your manuscript
from the markdown file and was only able to look at the version already in your directory. I got this error: File ../../assets/american-journal-of-epidemiology.csl not found in resource path

### Summary assessment

* small parts not reproducible or required manual intervention

## Thoroughness
How thorough was the overall study? Were alternatives (e.g. different ways of processing the data or different models) considered? Were alternatives discussed? Were the questions/hypotheses fully and thoroughly addressed?

### Feedback and Comments

As I said above, I think you were very thorough.

### Summary assessment

* strong level of thorougness

## Further comments

Make sure that all scientific names are italicized, both in the manuscript text and in the figures. For the figures, you can use the ggmarkdown package. Overall, I think your project was very well done and your manuscript is coming along nicely. Great job!

### 2025-04-26 products/manuscript/Manuscript.qmd

--- 2025-04-21 products/manuscript/Manuscript.qmd
@@
 +:==========+==========:+==========:+==========:+==========:+
-| (         | -2 3 2    | 1 57      | 1         | 0.1729390 |
-| I         | 8.5151515 | 3.052665  | .48025251 |           |
-| ntercept) |           |           |           |           |
+| ( I       | -2 3 2    | 1 57      | 1         | 0.1729390 |
+| ntercept) | 8.5151515 | 3.052665  | .48025251 |           |
 +-----------+-----------+-----------+-----------+-----------+
@@
 | EHRLICHIA | 5.0000000 | 5.157889  | 0         | 0.3576794 |
-| C         |           |           | .96938889 |           |
-| H         |           |           |           |           |
+| C H       |           |           | .96938889 |           |
 | AFFEENSIS |           |           |           |           |
@@
 +--------+--------+--------+--------+--------+
-| term   | e      | s t    | s t    | p      |
-|        | s      | d      | a      | .value |
-| \      | timate | .error | tistic |        |
-| <chr\> |        |        |        | \      |
-|        | \      | \      | \      | <dbl\> |
-|        | <dbl\> | <dbl\> | <dbl\> |        |
+| term   | e s    | s t d  | s t a  | p      |
+|        | timate | .error | tistic | .value |
+| \      |        |        |        |        |
+| \      | \      | \      | \      | \      |
+| <chr\> | \      | \      | \      | \      |
+|        | <dbl\> | <dbl\> | <dbl\> | <dbl\> |
 +:=======+=======:+=======:+=======:+=======:+
-| ( I n  | 9 9 3  | 5 7 2  | 1      | 0 .    |
-| t      | 5      | 9      | .      | 1      |
-| e      | .      | .      | 733920 | 434648 |
+| ( I n  | 9 9 3  | 5 7 2  | 1 .    | 0 . 1  |
+| t e    | 5 .    | 9 .    | 733920 | 434648 |
 | rcept) | 267857 | 946482 |        |        |
 +--------+--------+--------+--------+--------+
-| YEAR   | -   4  | 2      | -   1  | 0 .    |
-|        |     .  | .      |     .  | 1      |
-|        | 910714 | 835202 | 732051 | 438108 |
+| YEAR   | -   4  | 2 .    | -   1  | 0 . 1  |
+|        |     .  | 835202 |     .  | 438108 |
+|        |        |        |        |        |
+|        | 910714 |        | 732051 |        |
 +--------+--------+--------+--------+--------+

### 2025-05-03 products/manuscript/Manuscript.qmd
Final Project Push

@@
     highlight-style: github
-bibliography: ../../assets/dataanalysis-references.bib
-csl: ../../assets/american-journal-of-epidemiology.csl
+bibliography: ../../assets/references.bib
+csl: ../../assets/vancouver.csl
 editor: 
@@
 analysis involved extensive data cleaning to retain only confirmed and
-positive cases, with additional processing to resolve inconsistencies in
-timelines and reporting. Exploratory data analysis (EDA) was used to
-examine disease distribution, demographic patterns, geographic hotspots,
-and data completeness over time. Building on these insights, predictive
-models were developed to assess whether patient and case-level
-factors—such as age, disease type, and county—could help identify cases
-at risk of delayed follow-up. This work theorectically supports improved
-understanding of disease surveillance data and highlights areas for
-strengthening tick-borne disease reporting and case management systems
-in Georgia.
+positive cases, with a\@perumalsamy2024dditional processing to resolve
+inconsistencies in timelines and reporting. Exploratory data analysis
+(EDA) was used to examine disease distribution, demographic patterns,
+geographic hotspots, and data completeness over time. Building on these
+insights, predictive models were developed to assess whether patient and
+case-level factors—such as age, disease type, and county—could help
+identify cases at risk of delayed follow-up. This work theorectically
+supports improved understanding of disease surveillance data and
+highlights areas for strengthening tick-borne disease reporting and case
+management systems in Georgia.
 
@@
 
-### 2.1 Background
+### Background
 
@@
 greater public health concern, and expanded surveillance efforts across
-the globe. Researchers have documented the ecology and behavior of
-ticks, explored the role of climate and land use in tick habitat
-expansion, and analyzed patterns of disease incidence in both human and
-animal populations. Despite the growing body of research, many knowledge
-gaps remain, particularly in understanding co-infections, long-term
-health effects, and effective prevention strategies. Tick-borne diseases
-(TBDs) are primarily transmitted through the bite of infected
-hard-bodied ticks (family Ixodidae), and include bacterial, viral, and
-protozoan pathogens. Today, TBDs are recognized as a major public health
-concern in both rural and suburban settings, with transmission risk
-influenced by ecological, climatic, and sociocultural factors.
+the globe [@mcpherson2017]. Researchers have documented the ecology and
+behavior of ticks, explored the role of climate and land use in tick
+habitat expansion, and analyzed patterns of disease incidence in both
+human and animal populations. Despite the growing body of research, many
+knowledge gaps remain, particularly in understanding co-infections,
+long-term health effects, and effective prevention strategies
+[@usepa2016]. Tick-borne diseases (TBDs) are primarily transmitted
+through the bite of infected hard-bodied ticks (family Ixodidae), and
+include bacterial, viral, and protozoan pathogens. Today, TBDs are
+recognized as a major public health concern in both rural and suburban
+settings, with transmission risk influenced by ecological, climatic, and
+sociocultural factors [@perumalsamy2024].
 
-### 2.1.1 General Background Information
+### General Background Information
 
@@
 health concern, especially as the geographic range and seasonality of
-ticks expand due to environmental and ecological changes.
+ticks expand due to environmental and ecological changes @cdc2024a.
 
-### 2.1.2 Transmission and Reservoirs
+### Transmission and Reservoirs
 
@@
 themselves are influenced by multiple environmental factors, including
-temperature, humidity, vegetation, and host availability. Human exposure
-risk is increased in areas with high host density, edge habitats between
-forest and residential zones, and during seasons with peak tick
-activity. Additionally, land use changes—such as suburban development,
+temperature, humidity, vegetation, and host availability[@berger2014;
+@bouchard2019]. Human exposure risk is increased in areas with high host
+density, edge habitats between forest and residential zones, and during
+seasons with peak tick activity [@lerman2018; @aronson2017].
+Additionally, land use changes—such as suburban development,
 deforestation, and agricultural expansion—can modify tick habitats and
-affect pathogen dynamics.
+affect pathogen dynamics @allan2003.
 
-### 2.1.3 Ecological and Health Impacts
+### Ecological and Health Impacts
 
@@
 parasitism can influence population dynamics, particularly in rodents
-and ungulates. From a public health perspective, TBDs can cause a range
-of acute and chronic symptoms in humans, from fever and fatigue to
-long-term neurological and musculoskeletal complications.
-Co-infections—when an individual is infected with more than one
-pathogen—are increasingly recognized and may complicate diagnosis and
-treatment. The public health burden is also shaped by disparities in
-access to health care, awareness of TBDs, and availability of diagnostic
-testing, which may result in underreporting or misdiagnosis. The costs
-associated with tick-borne disease—including medical care, productivity
-loss, and prevention efforts—represent a growing concern for health
-systems, particularly in endemic areas.
+[@kasari2010; @clark]. From a public health perspective, TBDs can cause
+a range of acute and chronic symptoms in humans, from fever and fatigue
+to long-term neurological and musculoskeletal complications
+@zychowski2024. Co-infections—when an individual is infected with more
+than one pathogen—are increasingly recognized and may complicate
+diagnosis and treatment. The public health burden is also shaped by
+disparities in access to health care, awareness of TBDs, and
+availability of diagnostic testing, which may result in underreporting
+or misdiagnosis. The costs associated with tick-borne disease—including
+medical care, productivity loss, and prevention efforts—represent a
+growing concern for health systems, particularly in endemic areas
+@berghoff2012.
 
-### 2.1.4 Relevance
+### Relevance
 
@@
 increasing the geographic footprint and seasonality of tick-borne
-disease transmission. In the southeastern U.S., for example, Amblyomma
-americanum (the lone star tick) has expanded in abundance and geographic
-range over the past decades, contributing to increased cases of
-ehrlichiosis and the emergence of Alpha-Gal Syndrome, a red meat allergy
-triggered by tick bites. Understanding the ecology, distribution, and
-drivers of tick-borne disease is therefore crucial to predicting future
-trends, identifying at-risk populations, and developing effective
-prevention and response strategies.
+disease transmission [@gage2008; @dawe2016]. In the southeastern U.S.,
+for example, Amblyomma americanum (the lone star tick) has expanded in
+abundance and geographic range over the past decades, contributing to
+increased cases of ehrlichiosis and the emergence of Alpha-Gal Syndrome,
+a red meat allergy triggered by tick bites @cdc2022. Understanding the
+ecology, distribution, and drivers of tick-borne disease is therefore
+crucial to predicting future trends, identifying at-risk populations,
+and developing effective prevention and response strategies.
 
-# 3. The Data
+# The Data
 
-# 3.1 Data Acquisition & Description
+# Data Acquisition & Description
 
@@
 into the trends, distribution, and characteristics of reported cases,
-supporting public health efforts in disease monitoring and
-prevention.2.1.5 Description of Data and Data Source
+supporting public health efforts in disease monitoring and prevention.
 
@@
 
-## 3.1.2 Questions and Hypotheses to Be Addressed
+## Questions and Hypotheses to Be Addressed
 
@@
 
-# 4. Methods
+# Methods
 
-## 4.1 Workflow Summarized
+## Workflow Summarized
 
@@
 
-## 4.2 Uncleaned Data
+## Uncleaned Data
 
@@
 
-## 4.3 Initial Data Cleaning
+## Initial Data Cleaning
 
@@
 
-#### 4.3.1 Looked for Missing Data & General Trends
+#### Looked for Missing Data & General Trends
 
@@
 
-## 4.4 Exploratory/Descriptive Analysis
+## Exploratory/Descriptive Analysis
 
-#### 4.4.1 Identified Disease Burden Trends & Hotspot Counties
+#### Identified Disease Burden Trends & Hotspot Counties
 
@@
 
-#### 4.4.2 Checked Age Distribution
+#### Checked Age Distribution
 
@@
 
-#### 4.4.3 Evaluated Data Completeness Over Time
+#### Evaluated Data Completeness Over Time
 
@@
 
-#### 4.4.4 Analyzed Timeline Between Key Case Events
+#### Analyzed Timeline Between Key Case Events
 
@@
 
-## 4.5 EDA Revelant Takeaways
+## EDA Relevant Takeaways
 
-#### 4.5.1 Most Common Tick-Borne Diseases
+#### Most Common Tick-Borne Diseases
 
-The dataset is filtered to identify the five most frequently reported
+The data set is filtered to identify the five most frequently reported
 tick-borne diseases. A bar chart visualizes their case counts, offering
@@
 
-#### 4.5.2 Geographic Distribution
+**Figure 1.** Distribution of tick-borne disease case counts in Georgia,
+2019–2024.
+
+#### Geographic Distribution
 
@@
 
-#### 4.5.3 Data Completeness Over Time
+**Figure 2-4.** The distribution of the most common disease types across
+all counties– narrowed to the counties with more tick cases (hotspot
+counties) then further narrowed focus on the top five tick-borne
+diseases within these hotspot counties
+
+#### Data Completeness Over Time
 
@@
 
+**Figure 5.** Data completeness variation in the data set and entry
+consistency over time
+
 For more of my initial exploratory data analysis, please see the
@@
 
-#### 4.5.4 Timeline Analysis of Key Case Events
+#### Timeline Analysis of Key Case Events
 
@@
 
-## 4.6 Statistical Analysis
+## Statistical Analysis
 
@@
 
-# 5. Results
+# Results
 
-## 5.1 Basic statistical analysis
+## Basic statistical analysis
 
@@
 
-+-----------+-----------+-----------+-----------+-----------+
-| term      | estimate  | s         | statistic | p.value   |
-|           |           | td.error  |           |           |
-| \<chr\>   | \<dbl\>   |           | \<dbl\>   | \<dbl\>   |
-|           |           | \<dbl\>   |           |           |
-+:==========+==========:+==========:+==========:+==========:+
-| ( I       | -2 3 2    | 1 57      | 1         | 0.1729390 |
-| ntercept) | 8.5151515 | 3.052665  | .48025251 |           |
-+-----------+-----------+-----------+-----------+-----------+
-| YEAR      | 1.1515152 | 0.777581  | 1         | 0.1727716 |
-|           |           |           | .48089418 |           |
-+-----------+-----------+-----------+-----------+-----------+
-| B         | 0.4242424 | 5.484565  | 0         | 0.9400358 |
-| ABESIOSIS |           |           | .07735207 |           |
-+-----------+-----------+-----------+-----------+-----------+
-| DENGUE    | 4.8484848 | 6.364775  | 0         | 0.4656907 |
-|           |           |           | .76176850 |           |
-+-----------+-----------+-----------+-----------+-----------+
-| EHRLICHIA | 5.0000000 | 5.157889  | 0         | 0.3576794 |
-| C H       |           |           | .96938889 |           |
-| AFFEENSIS |           |           |           |           |
-+-----------+-----------+-----------+-----------+-----------+
-| EHRLICHIA | -         | 6.364775  | 0         | 0.8604390 |
-| EWINGII   | 1.1515152 |           | .18092002 |           |
-+-----------+-----------+-----------+-----------+-----------+
-| LYME      | 8.5606061 | 4.963745  | 1         | 0.1186814 |
-| DISEASE   |           |           | .72462659 |           |
-+-----------+-----------+-----------+-----------+-----------+
-| ROCKY     | 0.5757576 | 5.484565  | 0         | 0.9186958 |
-| MOUNTAIN  |           |           | .10497781 |           |
-| SPOTTED   |           |           |           |           |
-| FEVER     |           |           |           |           |
-+-----------+-----------+-----------+-----------+-----------+
-| WEST NILE | 8.2272727 | 5.593721  | 1         | 0.1754208 |
-| (WNV)     |           |           | .47080497 |           |
++----------------+-------------+-----------+-----------+-----------+
+| term\          | estimate\   | std.      | st        | p.value\  |
+| \<chr\>        | \<dbl\>     | error\    | atistic\  | \< dbl\>  |
+|                |             | \<dbl\>   | \<dbl\>   |           |
++:===============+============:+==========:+==========:+==========:+
+| (Intercept)    | \-          | 157       | 1.        | 0         |
+|                | 2           | 3.052665  |  48025251 | .1729390  |
+|                | 328.5151515 |           |           |           |
++----------------+-------------+-----------+-----------+-----------+
+| YEAR           | 1.1515152   | 0.777581  | 1.        | 0         |
+|                |             |           |  48089418 | .1727716  |
++----------------+-------------+-----------+-----------+-----------+
+| BABESIOSIS     | 0.4242424   | 5 .484565 | 0 .       | 0         |
+|                |             |           | 07735207  | .9400358  |
++----------------+-------------+-----------+-----------+-----------+
+| DENGUE         | 4.8484848   | 6\.       | 0 .       | 0         |
+|                |             | 364775    | 76176850  | .4656907  |
++----------------+-------------+-----------+-----------+-----------+
+| EHRLICHIA      | 5.0000000   | 5.157889  | 0 .       | 0         |
+| CHAFFEENSIS    |             |           | 96938889  | .3576794  |
++----------------+-------------+-----------+-----------+-----------+
+| EHRLICHIA      | \- 1 .      | 6.364775  | 0 .       | 0         |
+| EWINGII        | 1515152     |           | 18092002  | .8604390  |
++----------------+-------------+-----------+-----------+-----------+
+| LYME DISEASE   | 8.5606061   | 4.963745  | 1 .       | 0         |
+|                |             |           | 72462659  | .1186814  |
++----------------+-------------+-----------+-----------+-----------+
+| ROCKY MOUNTAIN | 0.5757576   | 5.484565  | 0\.       | 0         |
+| SPOTTED FEVER  |             |           | 10497781  | .9186958  |
++----------------+-------------+-----------+-----------+-----------+
+| WEST NILE      | 8.2272727   | 5.593721  | 1.        | 0         |
+| (WNV)          |             |           |  47080497 | .1754208  |
 | INFECTION      |             |           |           |           |
-+-----------+-----------+-----------+-----------+-----------+
++----------------+-------------+-----------+-----------+-----------+
 
 9 rows
+
+**Table 1.** Output form a linear model predicting total case counts by
+year and disease type. The intercept represents baseline case counts
+while each diseases represents additional effect size.
 
@@
 
-+--------+--------+--------+--------+--------+
-| term   | e s    | s t d  | s t a  | p      |
-|        | timate | .error | tistic | .value |
++---------+---------+---------+---------+---------+
+| term    | e       | std .   | s t a t | p .     |
+|         | stimate | error   | istic   | value   |
 | \       |         |         |         |         |
 | \       | \       | \       | \       | \       |
-| <chr\> | \      | \      | \      | \      |
-|        | <dbl\> | <dbl\> | <dbl\> | <dbl\> |
-+:=======+=======:+=======:+=======:+=======:+
-| ( I n  | 9 9 3  | 5 7 2  | 1 .    | 0 . 1  |
-| t e    | 5 .    | 9 .    | 733920 | 434648 |
-| rcept) | 267857 | 946482 |        |        |
-+--------+--------+--------+--------+--------+
-| YEAR   | -   4  | 2 .    | -   1  | 0 . 1  |
-|        |     .  | 835202 |     .  | 438108 |
-|        |        |        |        |        |
-|        | 910714 |        | 732051 |        |
-+--------+--------+--------+--------+--------+
+| \<chr\> | \       | \       | \       | \       |
+|         | \<      | \<      | \<      | \<      |
+|         | dbl\>   | dbl\>   | dbl\>   | dbl\>   |
++:========+========:+========:+========:+========:+
+| ( I n   | 9935 .  | 5729 .  | 1 .     | 0 .     |
+| t       | 267857  | 946482  | 733920  | 1434648 |
+| ercept) |         |         |         |         |
++---------+---------+---------+---------+---------+
+| YEAR    | 4 .     | 2 .     | 1 .     | 0 .     |
+|         | 910714  | 835202  | 732051  | 1438108 |
++---------+---------+---------+---------+---------+
 
@@
 
-## 5.2 Further Statistical Analysis
+## Further Statistical Analysis
 
@@
 
-## 5.3 Adding in Poverty Rate Data by County
+## Adding in Poverty Rate Data by County
 
@@
 
-+--------------+--------------+--------------+--------------+
-|              |              |              |              |
-+--------------+--------------+--------------+--------------+
-|              |              |              |              |
-+--------------+--------------+--------------+--------------+
-
 {{< pagebreak >}}
 
-# 6. Discussion
+# Discussion
 
-## 6.1 Summary and Interpretation
+## Summary and Interpretation
 
@@
 contain meaningful predictive information regarding follow-up delays in
-surveillance data.Such insights can inform improvements in case
-monitoring, triaging, and outreach in counties or patient groups with
-higher risk for delays.
+surveillance data. These results align with previous findings that
+individual-level demographics and geography can influence diagnosis of
+vector-borne diseases, though the moderate predictive power suggests
+additional factors— such as healthcare access and proximity
+[@starke2023; @frontier]— remain unmeasured in this simplistic modeled
+understanding of time to diagnosis. Such insights can inform
+improvements in case monitoring, triaging, and outreach in counties or
+patient groups with higher risk for delays.
 
@@
 of cases with diagnostic delays exceeding 30 days. A linear regression
-model indicated no statistically significant relationship between
-poverty rate and delay rate (p = 0.87). The estimated effect size was
-minimal (β = 0.0049), and the 95% confidence interval included zero.
+model indicated no statistically significant relationshipbetween poverty
+rate and delay rate (p = 0.87). The estimated effect size was minimal (β
+= 0.0049), and the 95% confidence interval included zero.
 
@@
 -   The all-predictors model performed more balanced, with a decent
-    accuracy and a stronger ROC AUC, meaning it did a better
-    job discriminating between delayed and non-delayed cases.
+    accuracy and a stronger ROC AUC, meaning it did a better job
+    discriminating between delayed and non-delayed cases.
 
@@
 
-## 6.2 Strengths and Limitations
+## Strengths and Limitations
 
@@
 
-# 7. Conclusions
+# Conclusions
 
